---
title: "Download AquaMatch Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sept 14, 2026"
---

# Background

This step downloads the AquaSat SDD and siteSR data and applies the Gardner-style
handoff coefficeints for use in a modeling workflow to estimate SDD across the 
intermountain west. Only run this script if there have been upstream changes
to the AquaMatch dataset, since downloading files takes some time to complete.

Knitting this file will not result in any download, as all chunks are set to
`eval = FALSE`.

## Setup

```{r libraries, eval=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(tidyverse)
```

```{r edi-login, eval=FALSE}
# login to EDI - this is a work around for error popping up in AquaMatchr
# requires EDIutils, which is required by AquaMatchr so no need to import
EDIutils::login(key = Sys.getenv('EDI_API_key'))
```

# Download AquaMatch data

Set a download location for the files. 

```{r aquamatch-dir, eval=FALSE}
# note location for file storage when downloading via AquaMatchr
aquamatch_dir <- "regional_clarity/aquamatch_files/downloads"
dir.create(aquamatch_dir, recursive = T)
```

Use the downlaod_parameters funciton to download the latest SDD files and cache
the file for later use.

```{r download-data, eval=FALSE}
# use AquaMatchr for data gathering
sdd <- download_parameters(parameters = "sdd", 
                           version = "newest")
# get it out of a list
sdd <- sdd[["sdd"]]
write_dataset(dataset = sdd, 
              path = file.path(aquamatch_dir, "sdd_cache.parquet"), 
              format = "parquet")
rm(sdd)
```

And now download the siteSR data.

```{r download-siteSR, eval=FALSE}
download_siteSR(save_location = aquamatch_dir, 
                version = "newest", 
                algal_mask = FALSE)
```

And build the stacked dataset.

```{r siteSR-stack, eval=FALSE}
build_sr(which_sr = "siteSR",
         sr_location = aquamatch_dir,
         algal_mask = FALSE,
         save = TRUE, 
         save_location = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.feather")
)

```

And also apply the Gardner-style handoff coefficients relative to Landsat 7.

```{r apply-handsoff, eval=FALSE}
# workaround for feather/parquet issue
open_dataset(file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.feather"), 
             format = "feather") %>% 
  write_dataset(dataset = .,  
                path = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.parquet"),
                format = "parquet")

# and apply handoffs
apply_handoffs(
  input_path = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.parquet"),
  handoff_path = list.files(aquamatch_dir, full.names = T)[grepl("handoff", list.files(aquamatch_dir))],
  correction_method = "Gardner_poly",
  sat_target = "LS7",
  algal_mask = "FALSE",
  save_location = "regional_clarity/aquamatch_files/downloads/siteSR_DSWE1_LS7corr.parquet"
)
```

Also download the scene-level metadata

```{r sitesr-scene, eval=FALSE}
download_sceneMetadata(save_location = aquamatch_dir, product = "siteSR")
```

