---
title: "Filter AquaMatch Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sept 14, 2026"
output:
  html_document
---

This step applies some filters to the SDD and siteSR data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(igraph)
library(data.table)
library(parallel)
library(arrow)
library(sf)
library(tigris)
library(tidyverse)

aquamatch_raw_data_dir <- "regional_clarity/aquamatch_files/downloads"
filtered_aquamatch_dir <- "regional_clarity/aquamatch_files"
```

## SDD data filtering

This section applies a ransac-style filter to sites with greater numbers of 
SDD observations to remove obvious outliers based on historical patterns and 
trends.

```{r}
sdd_raw_path <- file.path(aquamatch_raw_data_dir, "sdd_cache.parquet")

sdd <- open_dataset(sdd_raw_path)

sdd_filtered_path <- file.path(filtered_aquamatch_dir, "sdd_quality_filter.parquet")

# run a ransac-like algorithm per site to remove outliers
ransac_site <- function(date, one_site_sdd,
                        n_iter = 50,
                        mad_thresh = 2,
                        n_sd = 3,
                        range_buffer_frac = 0.15,
                        min_diff = 1,
                        min_sample = 8,
                        vote_frac = 0.6) {
  n <- length(one_site_sdd)
  if (n < min_sample + 1) return(rep(NA, n))  # too few points to run ransac
  
  t <- as.numeric(date)
  # tally, per point, how many valid iterations called it an inlier 
  inlier_votes <- rep(0L, n)
  n_valid_iter <- 0L
  
  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)
    out_idx <- setdiff(seq_len(n), idx)
    
    fit <- lm(one_site_sdd[idx] ~ t[idx])
    pred <- coef(fit)[1] + coef(fit)[2] * t
    resid <- one_site_sdd - pred
    # scale should come from held-out points only - points used to fit the
    # line have near-zero residual by construction, so including them
    # deflates the scale (and threshold) whenever min_sample is close to n,
    # making a single held-out point look like an outlier over a trivial
    # difference. use MAD rather than sd() here: sd() is itself inflated by
    # a cluster of genuine outliers landing in the held-out set (a handful of
    # large same-direction residuals drag the mean and blow up the squared
    # deviations), which loosens the threshold and lets that same cluster
    # vote itself in as inliers. MAD is robust to that as long as the
    # cluster is a minority of the held-out set, since it centers on the
    # median rather than the mean. 
    resid_scale <- if (length(out_idx) >= 2) {
      mad(resid[out_idx], constant = mad_thresh, na.rm = TRUE)
    } else {
      # too few held-out points (n == min_sample + 1) to trust their own MAD -
      # fall back to all-points MAD (deflated, but not degenerate) instead
      mad(resid, constant = mad_thresh, na.rm = TRUE)
    }
    # never flag a point closer than min_diff to the fit line, regardless of
    # how tight the scale-based threshold is - a < 1 m disagreement isn't
    # meaningfully an outlier for Secchi depth. 
    threshold <- max(n_sd * resid_scale, min_diff, na.rm = TRUE)
    inliers <- !is.na(resid) & abs(resid) < threshold
    
    inlier_votes <- inlier_votes + as.integer(inliers)
    n_valid_iter <- n_valid_iter + 1L
  }
  
  # indicate when no iteration ever produced a usable fit
  if (n_valid_iter == 0) return(rep(NA, n))
  
  # a point is an inlier if at least vote_frac of the valid iterations agreed
  best_inliers <- (inlier_votes / n_valid_iter) >= vote_frac
  
  # a flagged point can't be implausible for this site if its value falls
  # within the range other accepted (inlier) observations already span here -
  # deviating from the local trend line isn't grounds to flag it when the
  # site itself has already produced values in that range. this matters most
  # for low-n sites, where a short fit sample leaves little room to
  # distinguish a real trend from a value the site has simply shown before
  if (any(best_inliers)) {
    inlier_range <- range(sdd[best_inliers], na.rm = TRUE)
    # scale the buffer to this site's typical SDD rather than using a flat
    # meters value - 0.5 m is a huge margin at a 1 m-visibility site and
    # negligible at a 10 m-visibility site
    range_buffer <- range_buffer_frac * mean(sdd[best_inliers], na.rm = TRUE)
    best_inliers <- best_inliers | (sdd >= inlier_range[1] + range_buffer &
                                      sdd <= inlier_range[2] - range_buffer)
  }
  
  best_inliers
}


if (!file.exists(sdd_filtered_path)) {
  set.seed(43)
  sdd %>%
    filter(
      # do some SDD filters - must be in higher tiers of data
      tier < 3,
      # must have a reasonable  value
      harmonized_value >= 0.1,
      # if there are aggregated rows, they generally agree
      harmonized_value_cv <= 0.25 | is.na(harmonized_value_cv),
      # no flags
      misc_flag == 0,
      # not a bottom hit
      depth_flag == 0,
      # and we'll just grab lake/res/impoundment data
      ResolvedMonitoringLocationTypeName == "Lake, Reservoir, Impoundment"
    ) %>% 
    collect() %>% 
    arrange(harmonized_local_time)%>%
    mutate(is_inlier = ransac_site(harmonized_local_time, harmonized_value),
           .by = MonitoringLocationIdentifier) %>%
    # also reverse the flag in cases where multiple measurements are made
    # since we've already confirmed multiple measurements agree (value_cv <= 0.25)
    mutate(is_inlier = if_else(harmonized_row_count > 1 &
                                 (!is_inlier | is.na(is_inlier)), 
                               TRUE,
                               is_inlier)) %>%
    write_dataset(dataset = .,
                  path = sdd_filtered,
                  format = "parquet")
}
```

Check whether flagging concentrates at the SDD extremes rather than spreading
evenly across the distribution. A high-end concentration here isn't necessarily
a problem here: we expect a lot of what lands in
the far right tail of raw SDD is genuinely implausible (deeper than the
water body, deeper than the instrument's rope, transcription errors, etc.)
rather than real high-clarity variability - the point of this check is to
tell those two cases apart before trusting the filter.

```{r ransac-check-sdd-tails}
sdd_ransac_check <- open_dataset(sdd_filtered) %>% 
  collect()
```

Let's look at highly-flagged sites as a spot-check.

```{r}
high_flag_sites <- sdd_ransac_check %>% 
  summarize(n = n(), 
            n_flagged = sum(is_inlier == FALSE),
            mean_sdd = mean(harmonized_value),
            .by = c(MonitoringLocationIdentifier, lat, lon)) %>% 
  arrange(-n_flagged) %>% 
  filter(n_flagged >= 20)

map(.x = high_flag_sites$MonitoringLocationIdentifier,
    ~ sdd_ransac_check %>%
      filter(MonitoringLocationIdentifier == .x) %>%
      ggplot(., aes(x = ActivityStartDate,
                    y = harmonized_value,
                    color = is_inlier)) +
      geom_point() +
      labs(title = .x))
```

There are clearly some over and under flagging examples here, but it's a bit out
of scope to refine the ransac futher.

```{r}
sdd_ransac_check <- sdd_ransac_check %>% 
  mutate(is_inlier = case_when(MonitoringLocationIdentifier == "AZDEQ_SW-1293" &
                                 harmonized_value > 10 ~ FALSE,
                               MonitoringLocationIdentifier == "MNPCA-62-0024-00-202" & 
                                 ActivityStartDate < ymd("2005-01-01") ~ FALSE,
                               MonitoringLocationIdentifier == "WIDNR_WQX-643065" & 
                                 harmonized_value > 20 ~ FALSE,
                               .default = is_inlier))

write_feather(sdd_ransac_check, sdd_filtered)

```

Re-summarize and look again:

```{r}
high_flag_sites <- sdd_ransac_check %>% 
  summarize(n = n(), 
            n_flagged = sum(is_inlier == FALSE),
            mean_sdd = mean(harmonized_value),
            .by = c(MonitoringLocationIdentifier, lat, lon)) %>% 
  arrange(-n_flagged) %>% 
  filter(n_flagged >= 20)

map(.x = high_flag_sites$MonitoringLocationIdentifier,
    ~ sdd_ransac_check %>%
      filter(MonitoringLocationIdentifier == .x) %>%
      ggplot(., aes(x = ActivityStartDate,
                    y = harmonized_value,
                    color = is_inlier)) +
      geom_point() +
      labs(title = .x))
```

Let's also look to see what fraction of observations are indicated to be outliers:

```{r}
ggplot(sdd_ransac_check, aes(x = harmonized_value, fill = is_inlier)) +
  geom_histogram(position = "fill", bins = 30) +
  scale_fill_manual(values = c(`TRUE` = "grey70", `FALSE` = "red")) +
  labs(x = "observed Secchi (m)", y = "fraction of observations",
       fill = "inlier") +
  theme_bw()
```

This all looks okay.

## Remote Sensing Filters

Scene-level metadata (cloud cover, image quality, geometric accuracy, etc.)
isn't carried through siteSR's per-site aggregation, so we download it
separately and join it on below via `sat_id`.

```{r download-scene-metadata, message = FALSE, warning=FALSE}
# same interactive-overwrite-hang workaround as download-siteSR above
scene_metadata_paths <- list.files(
  aquamatch_raw_data_dir,
  pattern = "sceneMetadata_Landsat(457|89)\\.csv$",
  full.names = TRUE
)

# LS457 and LS89 have slightly different columns (e.g. a single
# IMAGE_QUALITY vs. split IMAGE_QUALITY_OLI/_TIRS), so bind_rows() unions
# them rather than assuming identical schemas
scene_metadata <- map(.x = scene_metadata_paths, 
                      .f = ~ open_csv_dataset(.) %>% 
                        filter(CLOUD_COVER < 20) %>% 
                        collect(), 
                      show_col_types = FALSE) %>%
  bind_rows()
```

We'll also do some site filtering - sites that are very close to 
shore and may have contamination should be removed from training.

```{r}
sites <- open_csv_dataset(file.path(aquamatch_raw_data_dir, "siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv")) %>%
  # make sure the point extent only touches one waterbody
  filter(number_int_wb == 1) %>%
  select(siteSR_id, wb_nhd_id, wb_gnis_name, wb_areasqkm,
         flag_thermal_TM_shoreline, flag_thermal_ETM_shoreline,
         flag_thermal_TIRS_shoreline, WGS84_Latitude, WGS84_Longitude) %>% 
  collect()

```

Different `siteSR_id` values (different source prefixes - `WQP_*`, `AM_*`,
`NWIS_*`, etc.) can still be the exact same physical monitoring location,
or two locations close enough together that Landsat can't tell them
apart. Two sites within 100m of each other but assigned to *different*
waterbodies aren't the same location - that's more likely two real,
distinct water features close together (or an NHD assignment error) than
one site measured twice - so sites are only collapsed together when
they're both within 100m of each other *and* share the same `wb_nhd_id`.
Grouping is transitive (connected components over that adjacency), so a
chain of near-duplicates collapses into a single `location_id` even if
its two most-distant members are themselves more than 100m apart.

```{r site-location-dedup}
site_pts <- sites %>%
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = 4326,
           remove = FALSE)

# sparse adjacency: which sites (including itself) fall within 100m of
# each site
site_within_100m <- st_is_within_distance(site_pts, dist = 100)

# ...then restrict those pairs to ones sharing a waterbody - equality is
# symmetric, so this keeps the adjacency symmetric (if j survives in i's
# filtered neighbor list, i survives in j's) for the connected-components
# step below
wb_nhd_id <- sites$wb_nhd_id
site_within_100m_same_wb <- Map(
  function(nbrs, wb) nbrs[wb_nhd_id[nbrs] == wb],
  site_within_100m, wb_nhd_id
)

# connected components over that adjacency - this is what makes the
# grouping transitive rather than strictly pairwise
site_components <- components(
  graph_from_adj_list(site_within_100m_same_wb, mode = "all")
)$membership

pad <- nchar(as.character(max(site_components)))
sites <- sites %>%
  mutate(location_id = sprintf(paste0("loc_%0", pad, "d"), site_components))

# one row per de-duplicated location, listing the individual siteSR_id
# values collapsed into it
site_location_groups <- sites %>%
  summarize(n_sites = n(),
            siteSR_ids = list(siteSR_id),
            WGS84_Latitude = mean(WGS84_Latitude),
            WGS84_Longitude = mean(WGS84_Longitude),
            .by = location_id) %>%
  arrange(-n_sites)

print(paste0(nrow(site_location_groups %>% filter(n_sites > 1)), " locations have 
             more than one site in the 100m area in the same waterbody"))

site_location_groups %>% 
  filter(n_sites > 20)

# for locations with more than one siteSR_id within 100m (see
# site_location_groups above), keep just the first siteSR_id per
# location_id, so a duplicated physical site doesn't feed the filter below
# more than once
representative_sites <- sites %>%
  distinct(location_id, .keep_all = TRUE) %>%
  pull(siteSR_id)

```


## Filter reflectance outliers before matching

AquaMatch's siteSR `unified_data` is 167.5 million rows across 342,704 sites nationwide, we
know that there are still outliers in the reflectance data, so we filter 
reflectance outliers here, *before* `match_siteSR_to_WQP()` ever sees the data, 
rather than after matching. We want to make sure
outlier filtering happens before matches are made so that a flagged bad RS
data point doesn't remove the matched row down with it. 

Instead of running on the full nationwide siteSR `unified_data`, we use
four scoping/QA steps to keep the expensive per-site filter below from
processing sites this project will never use or trusting scenes/sites it
shouldn't:

- **Regional pre-filter**: restrict to the HUC4 basins that actually
appear anywhere in this project's final 6-state regional output. 
- **Site de-duping**: The `location_id` dedup built alongside it
(collapsing same-waterbody sites within 100m) is used to identify one representative
site to process through the ransac filter
- **Metadata QA**: `pCount_dswe1/pCount_dswe_gt0 > 0.5`, `pCount_dswe1 > 8`,
`prop_clouds == 0` pretty generous filters to remove possbile contamination
- **Scene-level cloud QA**: `sat_id` is further restricted to scenes in
`scene_metadata` with
`CLOUD_COVER < 20`, catching a partly-cloudy scene that could have aerosol impacts
on the temp and optical data
- **Site geometry QA**: `siteSR_id` is restricted to the `sites` table
built above, which drops sites whose buffer overlaps with more than one waterbody
(`number_int_wb == 1`). 
site for the RANSAC-like filter.

Corrected bands, not raw bands, feed the ransac filter - reflectance isn't
comparable across missions without harmonizing first. 

The filter itself (`ransac_site_multiband()`) is the same per-site,
random-subsample RANSAC-like approach as `ransac_site()` above, just fitting
all seven corrected channels (six reflectance bands plus surface
temperature) per iteration instead of one, standardized by each channel's
own MAD-based scale from a single all-data fit: reflectance is a shakier fit
against a straight line over a site's full record than SDD is (blooms,
turbidity, and seasonality move bands around much more), so instead of a raw
residual threshold in reflectance units we standardize each band's
per-iteration residual by its own MAD and combine into one z-score per
point, then threshold on that combined score. 

The regression inside the iteration loop is written as closed-form OLS
rather than `lm()` to reduce processing time and we parallellize to further
optimize the time need here.

```{r siteSR-regional-band-ransac-setup}
siteSR_regional_filtered <- file.path(filtered_aquamatch_dir, "siteSR_DSWE1_regional_band_filtered.parquet")

# a second, diagnostics-only cache that retains all rows and inlier labels
# so later figures render on every knit without re-running the RANSAC fit
siteSR_regional_corr_diag <- file.path(aquamatch_raw_data_dir, "siteSR_DSWE1_regional_band_ransac_diagnostics.parquet")

# store the ransac funciton
ransac_site_multiband <- function(date, band_mat,
                                  n_iter = 100,
                                  n_mad = 2,
                                  # per-band sensor-noise floor
                                  min_diff = c(red_corr7 = 0.01, green_corr7 = 0.01,
                                               blue_corr7 = 0.01, nir_corr7 = 0.01,
                                               swir1_corr7 = 0.01, swir2_corr7 = 0.01,
                                               temp_corr7 = 1),
                                  min_sample = 8) {
  n <- nrow(band_mat)
  nb <- ncol(band_mat)
  if (n < min_sample + 1) return(rep(NA, n))  # too few points to run ransac
  
  # match by name rather than assuming column order lines up with min_diff
  min_diff_b <- min_diff[colnames(band_mat)]
  
  t <- as.numeric(date)
  
  # fixed per-band residual scale from a single all-data fit (MAD is robust
  # up to ~50% contamination, well above what we expect here) - reused for
  # every candidate subsample below instead of recomputed per iteration
  band_scale <- vapply(seq_len(nb), function(b) {
    y <- band_mat[, b]
    cf <- ols_fit(t, y)
    pred <- cf["intercept"] + cf["slope"] * t
    resid <- y - pred
    s <- mad(resid, na.rm = TRUE)
    if (is.na(s) || s == 0) s <- sd(resid, na.rm = TRUE)
    s
  }, numeric(1))
  
  best_inliers <- rep(FALSE, n)
  best_count <- 0
  
  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)
    
    # fit each band vs time on the *same* subsample, standardize each band's
    # residuals by its fixed scale, then combine - keeps one band's scale
    # from dominating and keeps the threshold comparable across candidate
    # lines of very different quality
    z <- matrix(NA_real_, nrow = n, ncol = nb)
    for (b in seq_len(nb)) {
      y <- band_mat[, b]
      if (length(unique(y[idx])) < 2 || is.na(band_scale[b]) || band_scale[b] == 0) next
      
      cf <- ols_fit(t[idx], y[idx])
      if (anyNA(cf)) next
      pred <- cf["intercept"] + cf["slope"] * t
      resid <- y - pred
      z[, b] <- abs(resid) / band_scale[b]
      # a raw difference this small is within sensor noise for this band -
      # never let it count as outlier evidence, regardless of how tight
      # band_scale is
      z[abs(resid) < min_diff_b[b], b] <- 0
    }
    
    combined_z <- rowMeans(z, na.rm = TRUE)
    inliers <- !is.na(combined_z) & combined_z < n_mad
    
    if (sum(inliers) > best_count) {
      best_count <- sum(inliers)
      best_inliers <- inliers
    }
  }
  
  # no iteration ever produced a usable fit - we haven't determined
  # anything about this site, so say so rather than leaving every point
  # flagged by the rep(FALSE, n) initialization
  if (best_count == 0) return(rep(NA, n))
  
  # a flagged point can't be implausible for this site if every one of its
  # band values falls within the range other accepted (inlier) points
  # already span for that band - mirrors the same rescue in ransac_site()
  if (any(best_inliers)) {
    in_range <- vapply(seq_len(nb), function(b) {
      y <- band_mat[, b]
      accepted <- y[best_inliers]
      if (length(accepted) == 0 || all(is.na(accepted))) return(rep(TRUE, n))
      rng <- range(accepted, na.rm = TRUE)
      y >= rng[1] & y <= rng[2]
    }, logical(n))
    best_inliers <- best_inliers | apply(in_range, 1, all)
  }
  
  best_inliers
}

# define closed-form OLS slope/intercept for y ~ t
ols_fit <- function(t, y) {
  ok <- is.finite(t) & is.finite(y)
  if (sum(ok) < 2 || length(unique(t[ok])) < 2) {
    return(c(intercept = NA_real_, slope = NA_real_))
  }
  tt <- t[ok]; yy <- y[ok]
  tm <- mean(tt); ym <- mean(yy)
  slope <- sum((tt - tm) * (yy - ym)) / sum((tt - tm)^2)
  c(intercept = ym - slope * tm, slope = slope)
}


# define the 6-state HUC2's
regional_huc2 <- ...

# define the bands
band_cols <- c("red_corr7", "green_corr7", "blue_corr7",
               "nir_corr7", "swir1_corr7", "swir2_corr7", "temp_corr7")


if (any(!file.exists(c(siteSR_regional_filtered, siteSR_regional_corr_diag)))) {
  
  # add some QA filters
  siteSR_regional_raw <- open_dataset(unified_data, format = "parquet") %>%
    filter(huc2 %in% regional_huc2,
           pCount_dswe1 / pCount_dswe_gt0 > 0.5,
           pCount_dswe1 > 8,
           prop_clouds == 0,
           # filter for scenes with < 20% cloud cover
           sat_id %in% scene_metadata$sat_id) 
  
  siteSR_regional_dedup <- siteSR_regional_raw %>% 
    fitler(# only choose one representative site per location
      siteSR_id %in% representative_sites
    ) %>%
    collect() 
  
  # run per site, parallelized across site-groups mc.cores leaves 2 cores free.
  # data.table (not dplyr group_by/mutate) for grouping
  set.seed(47)
  setDT(siteSR_regional_dedup)
  setorder(siteSR_regional_dedup, siteSR_id, date)
  
  site_splits <- split(siteSR_regional_dedup, by = "siteSR_id", keep.by = TRUE)
  
  site_splits_flagged <- mclapply(
    site_splits,
    function(site_df) {
      site_df[, is_band_inlier := ransac_site_multiband(date, as.matrix(.SD)),
              .SDcols = band_cols]
      site_df
    },
    mc.cores = max(1, detectCores() - 2)
  )
  
  # collapse and clean up
  siteSR_regional_corr <- rbindlist(site_splits_flagged)
  rm(site_splits, site_splits_flagged)
  gc()
  
  # siteSR_regional_corr (and its is_band_inlier flags) only covers
  # representative_sites - one siteSR_id per location_id. The other
  # siteSR_id(s) collapsed into each multi-site location_id (see
  # site-location-dedup above) are the same physical pixel(s), observed on
  # the same overpasses, so rather than re-running the expensive per-site
  # ransac_site_multiband() fit on them too, we reuse the representative
  # site's decision for that scene. sat_id is a scene-global id (Earth
  # Engine's system:index - see download_sceneMetadata.R), not a per-site
  # one, so the lookup must be keyed on (location_id, sat_id)
  location_lookup <- sites %>%
    select(siteSR_id, location_id)
  
  inlier_by_location_scene <- siteSR_regional_corr %>%
    select(siteSR_id, sat_id, is_band_inlier) %>%
    left_join(location_lookup, by = "siteSR_id") %>%
    select(location_id, sat_id, is_band_inlier) %>%
    distinct()
  
  # re-expand from "one representative row per location_id" back out to
  # every geometry-QA-passing siteSR_id, then attach each row's location_id's scene-level inlier
  # decision. A duplicate site's sat_id that never appeared for its
  # representative (or vice versa - e.g. one site's buffer had more
  # cloud/fewer clear pixels that scene) has no match here and comes through
  # as NA, which - like a ransac fit that never resolved - is dropped below
  # rather than assumed innocent.
  siteSR_regional_filtered <- siteSR_regional_raw %>%
    filter(siteSR_id %in% sites$siteSR_id) %>%
    left_join(location_lookup, by = "siteSR_id") %>%
    left_join(inlier_by_location_scene, by = c("location_id", "sat_id")) 
  
  # persist the pre-filter data (with is_band_inlier still attached) for the
  # diagnostics chunks below, since siteSR_regional_filtered_df further down
  # drops both the flagged rows and the is_band_inlier column itself
  write_dataset(siteSR_regional_filtered, siteSR_regional_corr_diag)
  
  siteSR_regional_filtered %>%
    filter(is_band_inlier | is.na(is_band_inlier)) %>%
    select(-location_id, -is_band_inlier) %>% 
    write_dataset(., siteSR_regional_filtered)
  
  rm(siteSR_regional_raw, siteSR_regional_corr, siteSR_regional_filtered_df,
     location_lookup, inlier_by_location_scene)
}
```

Look at some diagnostics to make sure this makes sense. These reload the
pre-filter data from `siteSR_regional_corr_diag` (written above) for diagnostics.

```{r load-siteSR-regional-corr-diag}
siteSR_regional_corr <- open_dataset(siteSR_regional_corr_diag) 
```

Pull sites where more than 10% of points were flagged, for a first look at
whether the filter is behaving reasonably.

```{r high-flag-sites}
high_flag_sr_sites <- siteSR_regional_corr %>%
  summarize(n = n(),
            n_flagged = sum(is_band_inlier == FALSE),
            .by = siteSR_id) %>%
  mutate(frac_flagged = n_flagged/n) %>%
  filter(frac_flagged > 0.1) %>%
  arrange(-frac_flagged) %>% 
  collect()
```

Spot-check a handful of sites with a large enough record (> 1500
observations) that the pattern is legible: flagged points should look like
genuine spectral excursions (residual cloud/shadow contamination, glint)
rather than a large fraction of any one site's record.

```{r spotcheck-sites}
spotcheck_sites <- siteSR_regional_corr[, .N, by = siteSR_id][N > 1500, siteSR_id]

## TODO: need to add location id here to identify shared location info
map(.x = spotcheck_sites,
    .f = ~ print(
      siteSR_regional_corr %>% 
        filter(siteSR_id == .x) %>%
        collect() %>% 
        pivot_longer(cols = all_of(band_cols),
                     names_to = "band",
                     values_to = "band_val") %>%
        ggplot(.,
               aes(x = date, y = band_val, color = is_band_inlier)) +
        geom_point() +
        facet_grid(band ~ siteSR_id, scales = "free_y") +
        scale_color_manual(values = c(`TRUE` = "grey40", `FALSE` = "red", NA = "grey10")) +
        theme_bw()
    )
)
```

Some of these look like they might belong to the same waterbody or are sites 
that are very close to one another, even after the de-duplication step.

Check whether flagging concentrates at reflectance extremes rather than
spreading evenly across the range of each band - the same "does this look
like real signal being cut, or generic scene noise" question asked of the
SDD-only filter above, just against the band values themselves since no
SDD/harmonized_value exists on this pre-match data. Even spreading here is
consistent with catching genuine spectral noise (cloud/shadow edges,
glint); concentration at one band's extremes would suggest the filter is
instead cutting real, legitimately bright or dark water.

```{r band-tail-check}
band_tail_check <- rbindlist(lapply(band_cols, function(b) {
  siteSR_regional_corr[, .(band = b, bin = ntile(get(b), 10), is_band_inlier)]
}))[, .(n = .N,
        n_flagged = sum(is_band_inlier == FALSE, na.rm = TRUE),
        n_unflagged = sum(is_band_inlier == TRUE, na.rm = TRUE),
        n_na = sum(is.na(is_band_inlier))), by = .(band, bin)
][, `:=`(frac_flagged = n_flagged / n,
         frac_unflagged = n_unflagged / n,
         frac_na = n_na / n)][order(band, bin)]

print(
  band_tail_check %>%
    pivot_longer(cols = c(frac_flagged, frac_unflagged, frac_na),
                 names_to = "status", values_to = "frac") %>%
    mutate(status = recode(status,
                           frac_flagged = "flagged",
                           frac_unflagged = "not flagged",
                           frac_na = "NA")) %>%
    ggplot(aes(x = bin, y = frac, fill = status)) +
    geom_col() +
    facet_wrap(~ band, scales = "free_x") +
    labs(x = "decile of corrected band value (1 = lowest)",
         y = "fraction of observations",
         fill = NULL) +
    theme_bw()
)
```

This does skew to flagging greater outliers at higher values, but that is probably
real. We can further look into that by looking at sites where there are a lot
of data being flagged in the 90th percentile.

Spot-check sites with a large enough record (> 1000 observations) that the
pattern is legible, restricted to sites with at least one point in that
band's decile 10 (i.e. at/above the band's global 90th percentile) - a
closer look at the sites/points actually behind the decile-10 concentration
surfaced above, to see whether those points look like isolated spectral
noise or a recurring signal (e.g. seasonal bloom) the filter is cutting.

```{r decile10-spotcheck-prep}
band_p90 <- setNames(
  vapply(band_cols, function(b) {
    quantile(siteSR_regional_corr[[b]], 0.9, na.rm = TRUE)
  }, numeric(1)),
  band_cols
)
band_p90_df <- data.table(band = band_cols, p90 = band_p90)

siteSR_band_long <- rbindlist(lapply(band_cols, function(b) {
  siteSR_regional_corr[, .(siteSR_id, date, band = b, band_val = get(b),
                           is_band_inlier)]
}))

large_sites <- siteSR_regional_corr[, .N, by = siteSR_id][N > 1000, siteSR_id]
decile10_sites_all <- siteSR_band_long[
  siteSR_id %in% large_sites & band_val >= band_p90[band], unique(siteSR_id)
]

# this condition flags the vast majority of large sites (a decile-10 hit in
# any of 7 bands is a low bar), so plotting all of them stops being a spot
# check - sample a fixed, reproducible subset instead
set.seed(43)
decile10_spotcheck_n <- 25
decile10_sites <- if (length(decile10_sites_all) > decile10_spotcheck_n) {
  sample(decile10_sites_all, decile10_spotcheck_n)
} else {
  decile10_sites_all
}
```

One figure per site, rather than one huge multi-site `facet_grid`, so each
site's timeseries is actually readable. `r length(decile10_sites_all)` sites
meet the decile10 criterion above; we spot-check a random sample of
`r length(decile10_sites)` of them here rather than rendering all of them.

```{r decile10-spotcheck-plot}
walk(decile10_sites, function(site) {
  print(
    siteSR_band_long[siteSR_id == site] %>%
      ggplot(aes(x = date, y = band_val, color = is_band_inlier)) +
      geom_point() +
      geom_hline(data = band_p90_df, aes(yintercept = p90),
                 linetype = "dashed") +
      facet_wrap(~ band, scales = "free_y", ncol = 2) +
      scale_color_manual(values = c(`TRUE` = "grey40", `FALSE` = "red")) +
      labs(title = site) +
      theme_bw()
  )
})
```

Can still see some related sites in here, but will leave as is for now.

```{r diagnostics-cleanup}
rm(siteSR_regional_corr, band_cols,
   high_flag_sr_sites, spotcheck_sites,
   band_tail_check, band_p90, band_p90_df,
   siteSR_band_long, large_sites,
   decile10_sites, decile10_sites_all, decile10_spotcheck_n)
gc()
```

