---
title: "Make Matchups"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Aug 11, 2026"
output: html_document
---

This code creates matchups between the AquaMatch SDD files and the siteSR
files. We use a generous 5-day window to work from for future steps.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(sf)
library(tmap)
library(AquaMatchr)
library(tidyverse)
```

```{r edi-login}
# login to EDI - this is a work around for error popping up in AquaMatchr
# requires EDIutils, which is required by AquaMatchr, so probably no issue here
EDIutils::login(key = Sys.getenv('EDI_API_key'))
```

## AquaMatch data

```{r aquamatch-dir}
# note location for file storage when downloading via AquaMatchr
aquamatch_dir <- "aquamatch_files"
dir.create(aquamatch_dir)
```

IMPORTANT: the chunk below must be run interactively in the R console at
least once (not via Knit) before this doc is knit. AquaMatchr's downloader
functions can prompt interactively (readline) to confirm overwriting
existing files, and knitting is non-interactive, so it hangs forever
waiting for an answer no one can give. Running it in the console first lets
you answer any prompt and populate the cache; once the cache files below
exist, knitting will just read from them and skip the network calls.

To force a fresh download (e.g. AquaMatch/siteSR released a new version),
delete the cached files from `aquamatch_files/` before re-running in the
console:
- `sdd_newest.rds`
- the `siteSR_Landsat*DSWE1_*.feather` files
- the `*sites_with_NHD_info*.csv` file
- the `lakeSR_collated_handoffs*.csv` file

```{r download-data, message = FALSE, warning=FALSE}
# download_parameters() doesn't cache to disk, so we cache it ourselves -
# otherwise every knit re-downloads from EDI
sdd_cache <- file.path(aquamatch_dir, "sdd_newest.rds")
if (file.exists(sdd_cache)) {
  sdd <- readRDS(sdd_cache)
} else {
  # use AquaMatchr for data gathering
  sdd <- download_parameters("sdd", "newest")
  # get it out of a list
  sdd <- sdd[[1]]
  saveRDS(sdd, sdd_cache)
}

# download_siteSR() writes files to aquamatch_dir, and if it finds files
# already there it prompts interactively (readline) to confirm overwrite -
# that prompt is what hangs the knit, since there's no console to answer it.
# so, check for existing files first and only download if they're missing
siteSR_existing <- list.files(
  aquamatch_dir,
  pattern = "siteSR_Landsat.*DSWE1_.*\\.feather$|sites_with_NHD_info.*\\.csv$|lakeSR_collated_handoffs.*\\.csv$",
  full.names = TRUE
)
if (length(siteSR_existing) > 0) {
  siteSR <- siteSR_existing
} else {
  siteSR <- download_siteSR(save_location = aquamatch_dir)
}
```

```{r sites}
# sites
sites <- read_csv(file.path(siteSR[grepl("sites_with_NHD_info", siteSR)]))

# just grab some info for SDD filtering and plotting
sites_less <- sites %>%
  select(siteSR_id,
         OrganizationIdentifier = org_id,
         MonitoringLocationIdentifier = loc_id)
```

```{r sdd-sites}
# just grab those which are visible
sdd_sites <- left_join(sites_less, sdd,
                       relationship = "many-to-many")

# reduce for matching
sdd_sites_less <- sdd_sites %>%
  select(siteSR_id, harmonized_utc, subgroup_id)

# do a quick summary
site_summary <- sdd_sites_less %>%
  summarise(n = n(), .by = siteSR_id)

# get unique site types here
unique(sdd_sites$ResolvedMonitoringLocationTypeName)
```

```{r lakes-only}
# let's limit to Lake data for our purposes
lakes_only <- sdd_sites %>%
  filter(grepl("Lake|Reservoir", MonitoringLocationTypeName)) %>%
  # remove the great lakes for this matching process
  filter(!grepl("Great", MonitoringLocationTypeName))

# and then reduce cols to match with sat data
lakes_for_sat <- lakes_only %>%
  select(siteSR_id, harmonized_utc, subgroup_id)
```

```{r lake-locs}
# look at those real quick
lake_locs <- sites %>%
  filter(siteSR_id %in% unique(lakes_for_sat$siteSR_id)) %>%
  st_as_sf(., coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = "EPSG:4326")

tmap_mode("plot")
tm_shape(lake_locs) +
  tm_dots()
```

```{r conus-only, , message = FALSE, warning=FALSE}
# let's drop the ones that are outside the CONUS for the purposes of this
states <- tigris::states() %>%
  filter(!grepl("AK|AS|MP|VI|HI|PR|GU", STUSPS))

lake_locs_CONUS <- lake_locs %>%
  st_transform(., st_crs(states)) %>%
  .[states,] %>%
  st_transform(., st_crs(lake_locs))
```

And plot

```{r}
tm_shape(lake_locs_CONUS) + tm_dots()
```

That's better, now let's reduce lakes for sat pairing with that info

```{r lakes-for-sat-conus}
lakes_for_sat <- lakes_for_sat %>%
  filter(siteSR_id %in% lake_locs_CONUS$siteSR_id)
```

## Make matchups

```{r make-matchups-function}
make_matchups <- function(sat_data_fp, wq_data) {
  # data
  wq_data <- wq_data %>%
    mutate(harmonized_utc = force_tz(harmonized_utc, "UTC"),
           wq_date = ymd(format(harmonized_utc, "%Y-%m-%d")))
  sat_data <- read_feather(sat_data_fp)  %>%
    filter(siteSR_id %in% unique(wq_data$siteSR_id))

  # reduce columns in satellite data
  sat_less <- sat_data %>%
    select(siteSR_id, sat_id,
           sat_date = date) %>%
    mutate(sat_date = ymd(sat_date))

  # coerce to data.table
  setDT(wq_data)
  setDT(sat_less)
  setDT(sat_data)

  # overjoin using data.table syntax (this is biiiiiiig)
  matches <- sat_less[
    wq_data,
    on = .(siteSR_id),
    allow.cartesian = TRUE
  ][
    # and filter to window of 5 days
    abs(as.integer(difftime(wq_date,
                            sat_date,
                            units = "days"))) <= 5
  ]

  # add the sat data back in
  left_join(matches, sat_data)
}
```

```{r apply-matchups}
# apply function
five_day_matches <- map2(.x = siteSR[grepl("feather", siteSR)],
                         .y = list(lakes_for_sat),
                         make_matchups) %>%
  bind_rows()

# join with SDD data
five_day_matches <- five_day_matches %>%
  left_join(., sdd)
```

```{r export-matches}
# export
write_feather(five_day_matches, "aquamatch_files/five_day_sdd_matches.feather")
```

## Summary and Figures

```{r summary-by-site}
# summarize by number of matches per site
five_day_summary <- five_day_matches %>%
  summarize(n = n(),
            .by = siteSR_id)

# add in spatial data
lake_locs_with_matches <- lake_locs_CONUS %>%
  filter(siteSR_id %in% unique(five_day_matches$siteSR_id)) %>%
  left_join(., five_day_summary)
```

```{r plot-all-missions}
# plot all missions
tm_shape(states) +
  tm_polygons() +
  tm_shape(lake_locs_with_matches, bbox = states) +
  tm_dots("n",
          fill.scale = tm_scale(breaks = c(1, 10, 25, 50, 75, 100,
                                           500, 1000, 1500),
                                values = "viridis"),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_layout(legend.position = tm_pos_in("right", "bottom"),
            legend.bg.color = "white",
            legend.frame = TRUE)
```

```{r summary-by-mission}
# and per site per mission
five_day_summary_sensor <- five_day_matches %>%
  summarize(n = n(),
            .by = c(siteSR_id, mission))

# add spatial data and label info
lake_locs_by_mission <- lake_locs_with_matches %>%
  select(siteSR_id, geometry) %>%
  right_join(five_day_summary_sensor, by = "siteSR_id") %>%
  mutate(mission = factor(mission,
                          levels = c("LT04", "LT05", "LE07", "LC08", "LC09"),
                          labels = c("Landsat 4", "Landsat 5", "Landsat 7", "Landsat 8", "Landsat 9")))%>%
  arrange(n)

# Compute totals per mission before faceting
mission_stats <- lake_locs_by_mission %>%
  st_drop_geometry() %>%
  summarize(total_matches = sum(n),
            total_sites = n_distinct(siteSR_id),
            .by = mission)

# and prep for viz
lake_locs_by_mission <- lake_locs_by_mission %>%
  left_join(mission_stats, by = "mission") %>%
  mutate(mission_label = paste0(mission, ", ",
                                format(total_matches, big.mark = ","), " matches, ",
                                format(total_sites, big.mark = ","), " sites")) %>%
  arrange(n)   # keep your draw-order fix from before
```

```{r plot-by-mission}
# plot!
tm_shape(states) +
  tm_polygons() +
  tm_shape(lake_locs_by_mission, bbox = states) +
  tm_dots("n",
          fill.scale = tm_scale(values = "viridis"),
          fill.legend = tm_legend(position = tm_pos_on_top(pos.h = "right", pos.v = "bottom"),
                                  bg.color = "white",
                                  frame = TRUE, text.size = 1),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_facets_wrap(by = "mission_label", ncol = 2, nrow = 3)
```
