---
title: "Make Matchups"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Aug 11, 2026"
output: html_document
---

This code creates matchups between the AquaMatch SDD files and the siteSR
files. We use a generous 5-day window to work from for future steps.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(sf)
library(tmap)
library(tigris)
library(AquaMatchr)
library(tidyverse)
```

```{r edi-login}
# login to EDI - this is a work around for error popping up in AquaMatchr
# requires EDIutils, which is required by AquaMatchr, so probably no issue here
EDIutils::login(key = Sys.getenv('EDI_API_key'))
```

## AquaMatch data

Download AquaMatch as necessary.

```{r aquamatch-dir}
# note location for file storage when downloading via AquaMatchr
aquamatch_dir <- "aquamatch_files"
dir.create(aquamatch_dir)
```

IMPORTANT: the chunk below must be run interactively in the R console at
least once (not via Knit) before this doc is knit. AquaMatchr's downloader
functions can prompt interactively (readline) to confirm overwriting
existing files, and knitting is non-interactive, so it hangs forever
waiting for an answer no one can give. Running it in the console first lets
you answer any prompt and populate the cache; once the cache files below
exist, knitting will just read from them and skip the network calls.

To force a fresh download (e.g. AquaMatch/siteSR released a new version),
delete the cached files from `aquamatch_files/` before re-running in the
console:
- `sdd_newest.rds`
- the `siteSR_Landsat*DSWE1_*.feather` files
- the `*sites_with_NHD_info*.csv` file
- the `lakeSR_collated_handoffs*.csv` file

```{r download-data, message = FALSE, warning=FALSE}
# download_parameters() doesn't cache to disk, so we cache it ourselves -
# otherwise every knit re-downloads from EDI
sdd_cache <- file.path(aquamatch_dir, "sdd_newest.feather")
if (!file.exists(sdd_cache)) {
  # use AquaMatchr for data gathering
  sdd <- download_parameters("sdd", "newest")
  # get it out of a list
  sdd <- sdd[[1]]
  write_feather(sdd, sdd_cache)
  rm(sdd)
}

# apply some QAQC filters for higher quality data and ransac filter
sdd_filtered <- file.path(aquamatch_dir, "sdd_quality_filter.feather")

# run a ransac-like algorithm per site to remove outliers
ransac_site <- function(date, sdd, n_iter = 200, threshold = 5, min_sample = 8) {
  n <- length(sdd)
  if (n < min_sample + 1) return(rep(FALSE, n))  # too few points to do anything
  
  t <- as.numeric(date)
  best_inliers <- rep(FALSE, n)
  best_count <- 0
  
  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)
    
    fit <- lm(sdd[idx] ~ t[idx])
    pred <- coef(fit)[1] + coef(fit)[2] * t
    resid <- abs(sdd - pred)
    inliers <- resid < threshold
    
    if (sum(inliers) > best_count) {
      best_count <- sum(inliers)
      best_inliers <- inliers
    }
  }
  best_inliers
}

if (!file.exists(sdd_filtered)) {
  read_feather(sdd_cache) %>% 
    filter(
      # do some SDD filters - must be in higher tiers of data
      tier < 3, 
      # must have a reasonable  value
      harmonized_value >= 0.1, 
      # no flags
      misc_flag == 0, 
      # not a bottom hit
      depth_flag == 0
    ) %>% 
    group_by(MonitoringLocationIdentifier) %>%
    arrange(harmonized_local_time) %>%
    mutate(is_inlier = ransac_site(harmonized_local_time, harmonized_value)) %>% 
    ungroup() %>% 
    write_feather(sdd_filtered)
}

# download_siteSR() writes files to aquamatch_dir, and if it finds files
# already there it prompts interactively (readline) to confirm overwrite -
# that prompt is what hangs the knit, since there's no console to answer it.
# so, check for existing files first and only download if they're missing
siteSR_existing <- list.files(
  aquamatch_dir,
  pattern = "siteSR_Landsat.*DSWE1_.*\\.feather$|sites_with_NHD_info.*\\.csv$|lakeSR_collated_handoffs.*\\.csv$",
  full.names = TRUE
)
if (length(siteSR_existing) > 0) {
  siteSR <- siteSR_existing
} else {
  siteSR <- download_siteSR(save_location = aquamatch_dir)
}

# stack the SR product
unified_data <- file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.feather")
if (!file.exists(unified_data)) {
  # stack SR product
  file_list <- siteSR[grepl(".feather", siteSR)]
  
  # Read files and stack - pulled from AquaMatchr buildSR() function
  # We use the {arrow} package to concatenate the datasets into a single table
  # instead of something like rbind(), which likely will use more memory than is
  # available to the user.
  unified_sr_dataset <- open_dataset(
    # All files indicated to contain SR data
    sources = file_list,
    # The files should not be assumed to have Hive-style partitioning
    hive_style = FALSE,
    # They are saved as .feather files
    format = "feather",
    # Don't assume all files in the file_list have the same scheme. For
    # example, Aerosols cols may not be present in all
    unify_schemas = TRUE
  ) %>%
    # Convert to Arrow Table
    as_arrow_table()
  
  write_feather(unified_sr_dataset, 
                file.path(aquamatch_dir, 
                          "siteSR_DSWE1_full_concatenation.feather"))
  
  # remove the dataset from workingspace
  rm(unified_sr_dataset)
}

# Make the matchups using AquaMatchr

# Define paths to downloaded and stacked data
sr_sites <- siteSR[grepl("collated_WQP_NWIS_sites", siteSR)]
out_file <- file.path(aquamatch_dir, "sdd_siteSR_matchups_5d.parquet")

if (file.exists(out_file)) {
  sdd_match <- read_parquet(out_file)
} else {
  # Work around a duckdb bug where it fails to create its temp directory
  # because the parent "duckdb" folder under tempdir() doesn't exist yet
  dir.create(file.path(tempdir(), "duckdb"), recursive = TRUE, showWarnings = FALSE)
  
  # Run the matchups with a 5-day window
  match_siteSR_to_WQP(
    wqp_path = sdd_filtered,
    siteSR_path = unified_data,
    site_list_path = sr_sites,
    save_location = out_file,
    time_window = "5 days"
  )
  
  # Read file
  sdd_match <- read_parquet(out_file)
}

```

Let's limit to Lake data for our purposes and remove the Great Lakes.

```{r lakes-only}
lake_sdd <- sdd_match %>%
  filter(
    # select for lake/reservoir only
    grepl("Lake|Reservoir", MonitoringLocationTypeName),
    # remove the great lakes for this matching process
    !grepl("Great", MonitoringLocationTypeName)
  )


rm(sdd_match)

write_feather(lake_sdd, file.path(aquamatch_dir, "sdd_CONUS_lake_5d_matchups.feather"))
```

Do some spatial filtering to CONUS only

```{r conus-only, , message = FALSE, warning=FALSE}
sites <- read_csv(sr_sites)
sites_w_matches <- sites %>% 
  filter(siteSR_id %in% unique(lake_sdd$siteSR_id)) %>% 
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = "EPSG:4326")

# let's drop the ones that are outside the CONUS for the purposes of this
conus <- suppressMessages(states(progress_bar = FALSE)) %>%
  filter(!grepl("AK|AS|MP|VI|HI|PR|GU", STUSPS))

lake_locs_CONUS <- sites_w_matches %>%
  st_transform(., st_crs(conus)) %>%
  .[conus,] %>%
  st_transform(., st_crs(sites_w_matches))
```

And plot

```{r}
tm_shape(conus) +
  tm_borders() +
  tm_shape(lake_locs_CONUS) + 
  tm_dots()
```


## Summary and Figures

Make some summary tables

```{r summary-by-site}
# summarize by number of matches per site
five_day_summary <- lake_sdd %>%
  summarize(n = n(),
            .by = siteSR_id)

# add in spatial data
lake_locs_with_matches <- lake_locs_CONUS %>%
  filter(siteSR_id %in% unique(lake_sdd$siteSR_id)) %>%
  left_join(., five_day_summary)
```

Plot locations colored by number of matches

```{r plot-all-missions}
# plot all missions
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_with_matches, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(breaks = c(1, 10, 25, 50, 75, 100,
                                           500, 1000, 1500),
                                values = "viridis"),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_layout(legend.position = tm_pos_in("right", "bottom"),
            legend.bg.color = "white",
            legend.frame = TRUE)
```

And also by number of matches per mission

```{r summary-by-mission}
# and per site per mission
five_day_summary_sensor <- lake_sdd %>%
  summarize(n = n(),
            .by = c(siteSR_id, mission))

# add spatial data and label info
lake_locs_by_mission <- lake_locs_CONUS %>%
  select(siteSR_id, geometry) %>%
  right_join(five_day_summary_sensor, by = "siteSR_id") %>%
  mutate(mission = factor(mission,
                          levels = c("LT04", "LT05", "LE07", "LC08", "LC09"),
                          labels = c("Landsat 4", "Landsat 5", "Landsat 7", "Landsat 8", "Landsat 9")))%>%
  arrange(n)

# Compute totals per mission before faceting
mission_stats <- lake_locs_by_mission %>%
  st_drop_geometry() %>%
  summarize(total_matches = sum(n),
            total_sites = n_distinct(siteSR_id),
            .by = mission)

# and prep for viz
lake_locs_by_mission <- lake_locs_by_mission %>%
  left_join(mission_stats, by = "mission") %>%
  mutate(mission_label = paste0(mission, ", ",
                                format(total_matches, big.mark = ","), " matches, ",
                                format(total_sites, big.mark = ","), " sites")) %>%
  arrange(n)   

# plot!
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_by_mission, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(values = "viridis"),
          fill.legend = tm_legend(position = tm_pos_on_top(pos.h = "right", pos.v = "bottom"),
                                  bg.color = "white",
                                  frame = TRUE, text.size = 1),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_facets_wrap(by = "mission_label", ncol = 2, nrow = 3)
```
