---
title: "Make Matchups"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sept 15, 2026"
output:
  html_document:
    fig_retina: 1
---

This code creates matchups between the filtered AquaMatch SDD and siteSR
files (from `01_filter_AquaMatch_Data.Rmd`) using a generous 5-day window.
Both inputs are already scoped to the HUC4 basins that make up this
project's 6-state regional AOI (that scoping happens in
`01_filter_AquaMatch_Data.Rmd`, before the expensive per-site filtering
there runs), so the "Subset to the regional AOI" section below is a
labeling/sanity-check step rather than an additional filter.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(sf)
library(tmap)
library(tigris)
library(tidyverse)

aquamatch_raw_data_dir <- "regional_clarity/aquamatch_files/downloads"
filtered_aquamatch_dir <- "regional_clarity/aquamatch_files"

# outputs from 01_filter_AquaMatch_Data.Rmd
sdd_filtered_path <- file.path(filtered_aquamatch_dir, "sdd_quality_filter.parquet")
siteSR_regional_filtered_path <- file.path(filtered_aquamatch_dir, "siteSR_DSWE1_regional_band_filtered.parquet")
sr_sites <- file.path(aquamatch_raw_data_dir, "siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv")
```

## Make matches with the filtered data

Use the regionally band-filtered siteSR data (rather than the full
nationwide `unified_data`) as the input to matching, so a site flagged in
script 01 simply isn't a match candidate.

`match_siteSR_to_WQP()` only accepts `.csv`/`.feather` for its WQP input and
hard-codes `.feather` for its siteSR input - parquet isn't a supported input
format for this AquaMatchr function. Our filtered caches from script 01 are
parquet (see that script's note on preferring parquet everywhere it isn't
required to be otherwise), so we write throwaway feather copies just for
this call and delete them immediately after.

```{r run-siteSR-wqp-match, message = FALSE, warning=FALSE}
out_file <- file.path(filtered_aquamatch_dir, "sdd_regional_siteSR_matchups_5d.parquet")

if (file.exists(out_file)) {
  sdd_match <- read_parquet(out_file)
} else {
  # throwaway feather copies, required by match_siteSR_to_WQP()'s input format
  sdd_filtered_feather <- tempfile(fileext = ".feather")
  siteSR_regional_filtered_feather <- tempfile(fileext = ".feather")
  # is_inlier == FALSE is a RANSAC-flagged SDD outlier (see
  # 01_filter_AquaMatch_Data.Rmd); NA (RANSAC never resolved, e.g. too few
  # points at that site) is kept rather than assumed guilty
  write_feather(read_parquet(sdd_filtered_path) %>%
                  filter(is_inlier | is.na(is_inlier)),
                sdd_filtered_feather)
  write_feather(read_parquet(siteSR_regional_filtered_path), siteSR_regional_filtered_feather)
  
  # Run the matchups with a 5-day window
  match_siteSR_to_WQP(
    wqp_path = sdd_filtered_feather,
    siteSR_path = siteSR_regional_filtered_feather,
    site_list_path = sr_sites,
    save_location = out_file,
    time_window = "5 days"
  )
  
  file.remove(sdd_filtered_feather, siteSR_regional_filtered_feather)
  
  sdd_match <- read_parquet(out_file)
}

```

Let's confirm we're only matching Lake data for our purposes and that the
Great Lakes were already excluded - this is a sanity check on the site
filters already applied upstream, not an additional filter.

```{r lakes-only}
lake_sdd <- sdd_match %>%
  filter(
    # select for lake/reservoir only
    grepl("Lake|Reservoir", MonitoringLocationTypeName),
    # remove the great lakes for this matching process
    !grepl("Great", MonitoringLocationTypeName)
  )

# check to see if these are the same
nrow(lake_sdd) == nrow(sdd_match)
```

## Summary Figures

```{r site-list, message = FALSE, warning=FALSE}
sites <- read_csv(sr_sites, show_col_types = FALSE)
```

```{r conus-only, , message = FALSE, warning=FALSE}
sites_w_matches <- sites %>%
  filter(siteSR_id %in% unique(sdd_match$siteSR_id)) %>%
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = "EPSG:4326")

# let's drop the ones that are outside the CONUS for the purposes of this
conus <- suppressMessages(states(progress_bar = FALSE)) %>%
  filter(!grepl("AK|AS|MP|VI|HI|PR|GU", STUSPS))

lake_locs_CONUS <- sites_w_matches %>%
  st_transform(., st_crs(conus)) %>%
  .[conus,] %>%
  st_transform(., st_crs(sites_w_matches))
```

And plot

```{r}
tm_shape(conus) +
  tm_borders() +
  tm_shape(lake_locs_CONUS) +
  tm_dots()
```

Make some summary tables

```{r summary-by-site}
# summarize by number of matches per site
five_day_summary <- sdd_match %>%
  summarize(n = n(),
            .by = siteSR_id)

# add in spatial data
lake_locs_with_matches <- lake_locs_CONUS %>%
  filter(siteSR_id %in% unique(sdd_match$siteSR_id)) %>%
  left_join(., five_day_summary)
```

Plot locations colored by number of matches

```{r plot-all-missions}
# plot all missions
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_with_matches, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(breaks = c(1, 10, 25, 50, 75, 100,
                                           500, 1000, 1500),
                                values = "viridis"),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_layout(legend.position = tm_pos_in("right", "bottom"),
            legend.bg.color = "white",
            legend.frame = TRUE)
```

And also by number of matches per mission

```{r summary-by-mission}
# and per site per mission
five_day_summary_sensor <- sdd_match %>%
  summarize(n = n(),
            .by = c(siteSR_id, mission))

# add spatial data and label info
lake_locs_by_mission <- lake_locs_CONUS %>%
  select(siteSR_id, geometry) %>%
  right_join(five_day_summary_sensor, by = "siteSR_id") %>%
  mutate(mission = factor(mission,
                          levels = c("LT04", "LT05", "LE07", "LC08", "LC09"),
                          labels = c("Landsat 4", "Landsat 5", "Landsat 7", "Landsat 8", "Landsat 9")))%>%
  arrange(n)

# Compute totals per mission before faceting
mission_stats <- lake_locs_by_mission %>%
  st_drop_geometry() %>%
  summarize(total_matches = sum(n),
            total_sites = n_distinct(siteSR_id),
            .by = mission)

# and prep for viz
lake_locs_by_mission <- lake_locs_by_mission %>%
  left_join(mission_stats, by = "mission") %>%
  mutate(mission_label = paste0(mission, ", ",
                                format(total_matches, big.mark = ","), " matches, ",
                                format(total_sites, big.mark = ","), " sites")) %>%
  arrange(n)

# plot!
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_by_mission, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(values = "viridis"),
          fill.legend = tm_legend(position = tm_pos_on_top(pos.h = "right", pos.v = "bottom"),
                                  bg.color = "white",
                                  frame = TRUE, text.size = 1),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_facets_wrap(by = "mission_label", ncol = 2, nrow = 3)
```

## Subset to the regional AOI

`siteSR` was already scoped to our 6-state region's HUC4 footprint upstream
in `01_filter_AquaMatch_Data.Rmd` (see that script's "Regional AOI" section)
- there's no regional-context-dependent processing between that scoping and
here, so every match in `sdd_match` should already fall inside it. We read
back the HUC4/state boundaries 01 already computed and wrote out, rather
than re-running the HUC4 download/overlap calculation, and use them here
only to label each site with its HUC4 (for the summaries below) and as a
sanity check on the upstream scoping.

```{r read-cached-boundaries, message = FALSE, warning=FALSE}
# standardize the geometry column name to "geometry" - the OpenFileGDB
# driver renames it to "SHAPE" on write, regardless of the name it had
# in 01_filter_AquaMatch_Data.Rmd before export
standardize_geom_col <- function(x) {
  geom_col <- attr(x, "sf_column")
  if (geom_col != "geometry") {
    names(x)[names(x) == geom_col] <- "geometry"
    st_geometry(x) <- "geometry"
  }
  x
}

HUC4_filtered <- st_read(file.path(filtered_aquamatch_dir, "huc4_filtered.gdb"),
                         layer = "huc4_filtered", quiet = TRUE) %>%
  standardize_geom_col()
states <- st_read(file.path(filtered_aquamatch_dir, "states.gdb"),
                  layer = "states", quiet = TRUE) %>%
  standardize_geom_col()
```

```{r huc4-label-join}
# add huc info - this is a labeling join (and a sanity check on the
# upstream regional scoping), not an additional filter: `left = FALSE` drops
# any matched site that somehow falls outside every regional HUC4, which
# shouldn't happen given the upstream scoping
regional_sites <- sites_w_matches %>%
  st_transform(st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered %>% select(huc4), left = FALSE)

cat(n_distinct(regional_sites$siteSR_id), "of",
    n_distinct(sites_w_matches$siteSR_id),
    "matched sites fall within the regional HUC4 footprint",
    "(should be all of them, since siteSR was already scoped to this",
    "footprint upstream in 01_filter_AquaMatch_Data.Rmd)\n")

# add sdd info back, carrying along each site's HUC4 label
regional_sdd <- sdd_match %>%
  filter(siteSR_id %in% regional_sites$siteSR_id) %>%
  left_join(
    regional_sites %>%
      st_drop_geometry() %>%
      select(siteSR_id, HUC4 = huc4) %>%
      # guard against a site sitting exactly on a shared HUC4 boundary,
      # which could otherwise match more than one polygon in the spatial join
      distinct(siteSR_id, .keep_all = TRUE),
    by = "siteSR_id"
  )
```

## QA the data

Make sure we're using the observation closest in time to the measurement.
This assures we're not using multiple satellites for the same data point in our
training dataset. We will also export those that are not included as a possible
additional test set.

```{r closest-obs}
# grab the observation closes in date to a satellite flyover
reg_sdd_closest <- regional_sdd %>%
  arrange(abs(time_diff), mission) %>%
  slice(1, .by = c("subgroup_id"))

# let's also do an anti-join here of the other observations so that we have an
# additional validation dataset to play with.
reg_sdd_additional <- anti_join(regional_sdd, reg_sdd_closest)

write_parquet(reg_sdd_additional,
              file.path(filtered_aquamatch_dir, "regional_sdd_not_closest.parquet"))
```

## Visualize data

Let's look at some longer records to do a sanity check to make sure that all
the filters worked correctly.

```{r site-summary}
# summarize by site for plotting
regional_site_summary <- reg_sdd_closest %>%
  st_as_sf(coords = c("lon", "lat"),
           crs = "EPSG:4326") %>%
  summarize(n = n(), .by = c(siteSR_id, geometry)) %>%
  arrange(n)

# look at some of the records with a lot of data
regional_long_record <- regional_site_summary %>%
  filter(n > 100) %>%
  left_join(., reg_sdd_closest)

ggplot(regional_long_record, aes(x = date, y = harmonized_value)) +
  geom_point() +
  facet_wrap(siteSR_id ~ .) +
  theme_bw()
```

Look at data spatially by site

```{r plot-sites-spatial}
tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  tm_shape(regional_site_summary %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_dots("n",
          fill.scale = tm_scale_continuous(values = "viridis"))
```

And summarize by HUC4

```{r plot-huc-summary}
regional_huc_summary <- reg_sdd_closest %>%
  st_as_sf(coords = c("lon", "lat"),
           crs = "EPSG:4326") %>%
  st_transform(., st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered, .) %>%
  summarize(n = sum(!is.na(siteSR_id)), .by = c(huc4, geometry))


tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  # drop 0-count HUC4s so they're left uncolored (outline only) instead of
  # taking the low end of the viridis scale
  tm_shape(regional_huc_summary %>% filter(n > 0)) +
  tm_polygons("n",
              fill.scale = tm_scale_continuous(values = "viridis"),
              fill_alpha = 0.5)

```

## Data Retention Summary

How much data survived each filtering step, from the raw nationwide SDD and
siteSR inputs down to the final matched, closest-observation training set?
`01_filter_AquaMatch_Data.Rmd` logged row/site counts through its SDD and
siteSR filtering steps (`row_loss_log_01.csv`); we append this script's
matching and closest-observation QA steps here. Every step is already
scoped to this project's 6-state HUC4 region except each pipeline's first
("raw") row.

```{r row-log-combine}
row_log <- read_csv(file.path(filtered_aquamatch_dir, "row_loss_log_01.csv"),
                    show_col_types = FALSE) %>%
  bind_rows(
    tibble(pipeline = "matched", step = "0_5day_window_match",
           n_rows = nrow(sdd_match), n_sites = n_distinct(sdd_match$siteSR_id)),
    tibble(pipeline = "matched", step = "1_closest_observation_final",
           n_rows = nrow(reg_sdd_closest), n_sites = n_distinct(reg_sdd_closest$siteSR_id))
  )

write_csv(row_log, file.path(filtered_aquamatch_dir, "row_loss_log.csv"))
```

```{r row-log-table}
row_log_display <- row_log %>%
  mutate(pipeline = factor(pipeline, levels = c("SDD", "siteSR", "matched"))) %>%
  arrange(pipeline, step) %>%
  group_by(pipeline) %>%
  mutate(step_label = step %>%
           str_remove("^[0-9]+[a-z]?_") %>%
           str_replace_all("_", " ") %>%
           str_to_sentence(),
         pct_of_raw = round(100 * n_rows / first(n_rows), 1),
         pct_of_previous = round(100 * n_rows / lag(n_rows), 1)) %>%
  ungroup() %>%
  select(pipeline, step_label, n_rows, n_sites, pct_of_previous, pct_of_raw)

knitr::kable(row_log_display,
             col.names = c("Pipeline", "Step", "Total Rows", "Total Sites",
                           "% of previous step retained", "% of pipeline's first step retained"),
             format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  # merge consecutive same-pipeline cells into one spanning cell, so the
  # table reads as three grouped blocks (SDD/siteSR/matched) rather than a
  # flat list repeating the pipeline name on every row
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

```{r row-log-figure, fig.height=8}
row_log_display %>%
  # fct_inorder() alone would put the first (largest) step at the bottom
  # after coord_flip() - reverse it so bars read largest-to-smallest top-down
  mutate(step_label = fct_rev(fct_inorder(step_label))) %>%
  ggplot(aes(x = step_label, y = n_rows)) +
  geom_col() +
  geom_text(aes(label = scales::comma(n_rows)), hjust = -0.05, size = 3) +
  facet_wrap(~ pipeline, scales = "free", ncol = 1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(x = NULL, y = "rows remaining",
       title = "Data retention through the filtering pipeline") +
  ggthemes::theme_few() +
  theme(legend.position = "none")
```

## Export data

```{r export}
write_parquet(reg_sdd_closest,
              file.path(filtered_aquamatch_dir, 
                        "filtered_regional_sdd.parquet"))
```
