---
title: "Subset to Region"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sept 01, 2026"
output: html_document
---

This code reduces the 5-day matchups to those in the HUC4's of a 6-state region
including Colorado, Wyoming, Utah, Idaho, Monana and New Mexico and
applies some QA filters to the data to create a machine learning dataset.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data prep

```{r libraries, message = FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(sf)
library(arrow)
library(tigris)
library(tmap)
library(hydrogeofetch)

# read in 5-day match
sdd_5d <- read_feather("aquamatch_files/sdd_CONUS_lake_5d_matchups.feather")
```

### Reduce AOI to Mountain West HUC4's

Get site info.

```{r unique-sites}
sites <- read_csv("aquamatch_files/siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv")

sites_filtered <- sites %>%
  filter(siteSR_id %in% unique(sdd_5d$siteSR_id)) %>%
  st_as_sf(., coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = "EPSG:4326")
```

Filter to states we care about.

```{r states-aoi, message = FALSE, warning=FALSE}
states <- states(progress_bar = FALSE) %>%
  filter(STUSPS %in% c("CO", "WY", "UT", "ID", "MT", "NM"))
states_union <- states %>% st_union()
```

Grab all the HUC4's from NHD and validate them

```{r huc4}
# get_huc() kept timing out (HTTP 504) against the USGS pygeoapi service, even
# retried state-by-state, so instead download the national WBD geodatabase
# once (cached in aquamatch_files/) and filter to our HUC4's locally.
wbd_gdb <- download_wbd(outdir = "aquamatch_files/wbd_national")

huc4_layer <- st_layers(wbd_gdb)$name %>%
  keep(~ grepl("HU4$", ., ignore.case = TRUE))

HUC4_national <- st_read(wbd_gdb, layer = huc4_layer, quiet = TRUE) %>%
  rename_with(tolower) %>%
  # WBD ships with Z values and a compound CRS (NAD83 + NAVD88 height);
  # transforming a 2D CRS into that compound CRS silently breaks (no vertical
  # grid to invent a Z), so drop Z here - horizontal datum is then plain NAD83
  st_zm(drop = TRUE, what = "ZM")

# standardize the geometry column name to "geometry" (FileGDB layers are
# often read in with the geometry column named "shape")
geom_col <- attr(HUC4_national, "sf_column")
if (geom_col != "geometry") {
  names(HUC4_national)[names(HUC4_national) == geom_col] <- "geometry"
  st_geometry(HUC4_national) <- "geometry"
}

# reset (not transform) the CRS now that only the horizontal NAD83 datum
# remains, matching states_union - no reprojection needed
st_crs(HUC4_national) <- st_crs(states_union)

# reduce from the national layer to just the HUC4's touching our 6-state AOI
HUC4 <- HUC4_national %>%
  st_make_valid() %>%
  st_filter(states_union)

HUC4v <- st_make_valid(HUC4)
```

And then assess overlapping area - here we want to inlcude HUCs that overlap
the 6-state area by at least 50%.

```{r huc4-overlap}
overlap_areas <- st_make_valid(HUC4v) %>%
  st_intersection(st_make_valid(states_union)) %>%
  mutate(overlap_area = st_area(geometry)) %>%
  st_drop_geometry() %>%
  select(huc4, overlap_area)

# calculate proportion overlap, filter for those greater than/equal to 50%
HUC4_filtered <- HUC4v %>%
  left_join(overlap_areas, by = "huc4") %>%
  mutate(
    overlap_area = coalesce(overlap_area, units::set_units(0, "m^2")),
    overlap_frac = as.numeric(overlap_area / st_area(geometry))
  ) %>%
  filter(overlap_frac >= 0.50)

# add huc info and filter to this extent
regional_sites <- sites_filtered %>%
  st_transform(st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered %>% select(huc4), left = FALSE)

# add sdd info back, carrying along each site's HUC4 label
regional_sdd <- sdd_5d %>%
  filter(siteSR_id %in% regional_sites$siteSR_id) %>%
  left_join(
    regional_sites %>%
      st_drop_geometry() %>%
      select(siteSR_id, HUC4 = huc4) %>%
      # guard against a site sitting exactly on a shared HUC4 boundary,
      # which could otherwise match more than one polygon in the spatial join
      distinct(siteSR_id, .keep_all = TRUE),
    by = "siteSR_id"
  )
```

## QA the data

Here, we apply some filters to make sure we are training the ML model with 
high-quality remote sensing data.

```{r qa-filters}
# apply some QA filters
reg_sdd_filtered <- regional_sdd %>%
  # RS data filter
  filter(
    # make sure at least 50% of pixels in area are water of some type
    pCount_dswe1/pCount_dswe_gt0 > 0.5,
    # make sure there are at least 8 pixels
    pCount_dswe1 > 8,
    # make sure there is no indication of cloud in the pixels
    prop_clouds == 0,
    # filter for inlier SDD only
    is_inlier
  )
```

And then making sure we're using the observation closest in time to the measurement.
This assures we're not using multiple satellites for the same data point in our 
training dataset. We will also export those that are not included as a possible
additional test set.

```{r closest-obs}
# grab the observation closes in date to a satellite flyover
reg_sdd_closest <- reg_sdd_filtered %>%
  arrange(abs(time_diff), mission) %>%
  slice(1, .by = c("subgroup_id"))

# let's also do an anti-join here of the other observations so that we have an
# additional validation dataset to play with.
reg_sdd_additional <- anti_join(reg_sdd_filtered, reg_sdd_closest)

write_feather(reg_sdd_additional, "aquamatch_files/regional_sdd_not_closest.feather")
```

## Visualize data

Let's look at some longer records to do a sanity check to make sure that all 
the filters worked correctly.

```{r site-summary}
# summarize by site for plotting
regional_site_summary <- reg_sdd_closest %>%
  st_as_sf(coords = c("lon", "lat"), 
           crs = "EPSG:4326") %>% 
  summarize(n = n(), .by = c(siteSR_id, geometry)) %>%
  arrange(n)

# look at some of the records with a lot of data
regional_long_record <- regional_site_summary %>%
  filter(n > 100) %>%
  left_join(., reg_sdd_closest)

ggplot(regional_long_record, aes(x = date, y = harmonized_value)) +
  geom_point() +
  facet_wrap(siteSR_id ~ .) +
  theme_bw()
```

Look at data spatially by site

```{r plot-sites-spatial}
tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  tm_shape(regional_site_summary %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_dots("n",
          fill.scale = tm_scale_continuous(values = "viridis"))
```

And summarize by HUC4

```{r plot-huc-summary}
regional_huc_summary <- reg_sdd_closest %>%
    st_as_sf(coords = c("lon", "lat"),
           crs = "EPSG:4326") %>%
  st_transform(., st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered, .) %>%
  summarize(n = sum(!is.na(siteSR_id)), .by = c(huc4, geometry))


tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  # drop 0-count HUC4s so they're left uncolored (outline only) instead of
  # taking the low end of the viridis scale
  tm_shape(regional_huc_summary %>% filter(n > 0)) +
  tm_polygons("n",
              fill.scale = tm_scale_continuous(values = "viridis"),
              fill_alpha = 0.5)

```

## Export data

```{r export}
write_feather(reg_sdd_closest,
              "aquamatch_files/filtered_regional_sdd.feather")
```
