---
title: "Split Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 1, 2026"
output: html_document
---

This code splits the filtered regional SDD dataset into 5 roughly-equal
parts, grouped by HUC4 so that a given HUC4 basin isn't divided across
splits, for downstream train/test/cross-validation use. We then map the
resulting split assignment as a sanity check.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message = FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(arrow)
library(sf)
library(tmap)
library(tigris)
library(hydrogeofetch)

# ROSSyndicate light palette for plotting
ROSS_lt_pal <- c("#002EA3", "#E70870", "#256BF5",
                 "#745CFB", "#1E4D2B", "#56104E")

# read in the filtered regional dataset from script 02
filtered_ml <- read_feather("aquamatch_files/filtered_regional_sdd.feather")
```

## Split into balanced partitions by HUC4

We want 5 partitions of roughly equal size, without splitting any single
HUC4's data across partitions. To do this, we sort HUC4s by row count
(largest first) and greedily assign each one to whichever of the 5 running
partitions currently has the fewest rows. This "greedy bin-packing" keeps
the partitions size-balanced while keeping each HUC4 intact within a single
partition.

```{r split-huc4}
total_rows <- nrow(filtered_ml)
target_size <- ceiling(total_rows / 5)

# row count per HUC4, largest first
huc4_sizes <- filtered_ml %>%
  group_by(HUC4) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# greedily assign each HUC4 to the currently-smallest partition
parts <- vector("list", 5)
part_sizes <- rep(0, 5)

for (i in 1:nrow(huc4_sizes)) {
  smallest_part <- which.min(part_sizes)
  parts[[smallest_part]] <- c(parts[[smallest_part]], huc4_sizes$HUC4[i])
  part_sizes[smallest_part] <- part_sizes[smallest_part] + huc4_sizes$count[i]
}

# subset df to the HUC4s in a given partition and label with the partition number
split_df <- function(df, parts, part_label) {
  map(parts, function(huc4_list) {
    df %>%
      filter(HUC4 %in% huc4_list)
  }) %>%
    bind_rows() %>%
    mutate(part = part_label)
}

# apply the split and label to all 5 partitions, then recombine
split_dataframes <- pmap(list(list(filtered_ml),
                              parts,
                              seq(1:5)),
                         split_df) %>%
  bind_rows()
```

## Visualize the split

Plot the resulting split assignment spatially as a check.

```{r map-splits}
# unique site locations, colored by split assignment
sites <- split_dataframes %>%
  summarize(n = n(), .by = c(siteSR_id, lon, lat, part)) %>%
  st_as_sf(.,
           coords = c("lon",
                      "lat"),
           crs = "EPSG:4326") %>%
  mutate(split = part)

# 6-state region and HUC4 boundaries for context
states <- states(progress_bar = FALSE) %>%
  filter(STUSPS %in% c("CO", "WY", "UT", "ID", "MT", "NM"))
states_union <- states %>% st_union()

# get_huc() (nhdplusTools/hydrogeofetch) times out (HTTP 504) against the USGS
# pygeoapi service, so instead read HUC4's from the national WBD geodatabase
# (downloaded once and cached in aquamatch_files/ by script 02) and filter
# to our 6-state AOI locally.
wbd_gdb <- download_wbd(outdir = "aquamatch_files/wbd_national")

huc4_layer <- st_layers(wbd_gdb)$name %>%
  keep(~ grepl("HU4$", ., ignore.case = TRUE))

HUC4s <- st_read(wbd_gdb, layer = huc4_layer, quiet = TRUE) %>%
  rename_with(tolower) %>%
  # WBD ships with Z values and a compound CRS (NAD83 + NAVD88 height);
  # transforming a 2D CRS into that compound CRS silently breaks (no vertical
  # grid to invent a Z), so drop Z here - horizontal datum is then plain NAD83
  st_zm(drop = TRUE, what = "ZM")

# standardize the geometry column name to "geometry" (FileGDB layers are
# often read in with the geometry column named "shape")
geom_col <- attr(HUC4s, "sf_column")
if (geom_col != "geometry") {
  names(HUC4s)[names(HUC4s) == geom_col] <- "geometry"
  st_geometry(HUC4s) <- "geometry"
}

# reset (not transform) the CRS now that only the horizontal NAD83 datum
# remains, matching states_union - no reprojection needed
st_crs(HUC4s) <- st_crs(states_union)

# reduce from the national layer to just the HUC4's touching our 6-state AOI
HUC4s <- HUC4s %>%
  st_make_valid() %>%
  st_filter(states_union)

tm_shape(HUC4s) +
  tm_shape(states) +
  tm_polygons(col = "grey70") +
  tm_borders(col = "grey20") +
  tm_shape(HUC4s) +
  tm_polygons(fill_alpha = 0) +
  tm_borders(col = "black") +
  tm_shape(sites) +
  tm_symbols(fill = "split",
             col = NA,
             size = 0.5,
             fill_alpha = 0.8,
             fill.scale = tm_scale_discrete(values = ROSS_lt_pal)) +
  tm_shape(states) +
  tm_text("STUSPS", col = "grey20") +
  tm_layout(frame = FALSE,
            legend.position = tm_pos_in("right", "top"))
```

## Export data

```{r export}
write_feather(split_dataframes, "aquamatch_files/split_filtered_regional.feather")
```
