---
title: "Split Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 1, 2026"
output: html_document
---

This code splits the filtered regional SDD dataset into 5 roughly-equal
parts, grouped by HUC4 so that a given HUC4 basin isn't divided across
splits, for downstream train/test/cross-validation use. We then map the
resulting split assignment as a sanity check.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message = FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(arrow)
library(sf)
library(tmap)

# ROSSyndicate light palette for plotting
ROSS_lt_pal <- c("#002EA3", "#E70870", "#256BF5",
                 "#745CFB", "#1E4D2B", "#56104E")

# read in the filtered regional dataset from script 02
filtered_ml <- read_feather("aquamatch_files/filtered_regional_sdd.feather")
```

## Split into balanced partitions by HUC4

We want 5 partitions of roughly equal size, without splitting any single
HUC4's data across partitions. To do this, we sort HUC4s by row count
(largest first) and greedily assign each one to whichever of the 5 running
partitions currently has the fewest rows. This "greedy bin-packing" keeps
the partitions size-balanced while keeping each HUC4 intact within a single
partition.

```{r split-huc4}
total_rows <- nrow(filtered_ml)
target_size <- ceiling(total_rows / 5)

# row count per HUC4, largest first
huc4_sizes <- filtered_ml %>%
  group_by(HUC4) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# greedily assign each HUC4 to the currently-smallest partition
parts <- vector("list", 5)
part_sizes <- rep(0, 5)

for (i in 1:nrow(huc4_sizes)) {
  smallest_part <- which.min(part_sizes)
  parts[[smallest_part]] <- c(parts[[smallest_part]], huc4_sizes$HUC4[i])
  part_sizes[smallest_part] <- part_sizes[smallest_part] + huc4_sizes$count[i]
}

# subset df to the HUC4s in a given partition and label with the partition number
split_df <- function(df, parts, part_label) {
  map(parts, function(huc4_list) {
    df %>%
      filter(HUC4 %in% huc4_list)
  }) %>%
    bind_rows() %>%
    mutate(part = part_label)
}

# apply the split and label to all 5 partitions, then recombine
split_dataframes <- pmap(list(list(filtered_ml),
                              parts,
                              seq(1:5)),
                         split_df) %>%
  bind_rows()
```

## Visualize the split

Plot the resulting split assignment spatially as a check.

```{r map-splits}
# unique site locations, colored by split assignment
sites <- split_dataframes %>%
  summarize(n = n(), .by = c(siteSR_id, lon, lat, part)) %>%
  st_as_sf(.,
           coords = c("lon",
                      "lat"),
           crs = "EPSG:4326") %>%
  mutate(split = part)

# 6-state region and HUC4 boundaries for context (saved by script 02, so we
# don't have to re-download/re-filter them here)
states <- st_read("aquamatch_files/states.gdb", layer = "states", quiet = TRUE)
HUC4s <- st_read("aquamatch_files/huc4_filtered.gdb", layer = "huc4_filtered",
                  quiet = TRUE)

# OpenFileGDB round-trips the geometry column as "SHAPE" - standardize back
# to "geometry"
geom_col <- attr(HUC4s, "sf_column")
if (geom_col != "geometry") {
  names(HUC4s)[names(HUC4s) == geom_col] <- "geometry"
  st_geometry(HUC4s) <- "geometry"
}

tm_shape(HUC4s) +
  tm_shape(states) +
  tm_polygons(col = "grey70") +
  tm_borders(col = "grey20") +
  tm_shape(HUC4s) +
  tm_polygons(fill_alpha = 0) +
  tm_borders(col = "black") +
  tm_shape(sites) +
  tm_symbols(fill = "split",
             col = NA,
             size = 0.5,
             fill_alpha = 0.8,
             fill.scale = tm_scale_discrete(values = ROSS_lt_pal)) +
  tm_shape(states) +
  tm_text("STUSPS", col = "grey20") +
  tm_layout(frame = FALSE,
            legend.position = tm_pos_in("right", "top"))
```

## Export data

```{r export}
write_feather(split_dataframes, "aquamatch_files/split_filtered_regional.feather")
```
