---
title: "Make Models"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 1, 2026"
output: html_document
---

This code trains XGBoost models to predict Secchi disk depth from Landsat
surface reflectance for the 6-state regional dataset. We apply the AquaMatch
cross-sensor corrections, engineer band-ratio spectral indices, split the
partitioned dataset from script 03 into cross-validation and test sets,
tune hyperparameters, train four leave-one-partition-out models, and
evaluate performance on the held-out test partition.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message = FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(arrow)
library(xgboost)
library(Metrics)
library(igraph)
library(sf)
library(tmap)
library(tigris)
library(hydrogeofetch)

# read in the partitioned dataset from script 03
split_dataframes <- read_feather("aquamatch_files/split_filtered_regional.feather")
```

## Apply cross-sensor corrections

Landsat missions differ slightly in their band responses, so we apply
Gardner et al.'s per-band quadratic correction coefficients (referenced to
Landsat 7) to harmonize reflectance values across missions before deriving
any indices.

```{r sensor-corrections}
am_corr_LS7 <- read_csv("aquamatch_files/lakeSR_collated_handoffs_GEEv2025-02-12_QAv2025-06-04.csv") %>%
  filter(correction == "Gardner", dswe == "DSWE1", sat_to == "LS7") %>% 
  # pivot for easier application
  pivot_longer(cols = c(intercept, B1, B2, min_in_val, max_in_val),
               names_to = "int_coef",
               values_to = "value") %>%
  mutate(new_column = paste(band, int_coef, sep = "_")) %>%
  select(-c(band, int_coef, slope, method)) %>%
  pivot_wider(names_from = new_column,
              values_from = value)

# apply coefficients to timeseries data
split_dataframes_corr <- split_dataframes %>%
  mutate(sat_corr = case_when(mission == "LT04" ~ "LS5", # landsat 4 is roughly the same as 5
                              mission == "LT05" ~ "LS5",
                              mission == "LE07" ~ "LS7",
                              mission == "LC08" ~ "LS8",
                              mission == "LC09" ~ "LS8", # landsat 9 is roughly the same as 8
                              TRUE ~ NA_character_),
         # LEDAPS atm correction for L4/5/7, LaSRC for L8/9
         atm_corr = case_when(mission %in% c("LT04", "LT05", "LE07") ~ "LEDAPS",
                              mission %in% c("LC08", "LC09") ~ "LaSRC",
                              TRUE ~ NA_character_),
         # numeric encoding for xgboost (LEDAPS as reference level)
         atm_corr_LaSRC = as.numeric(atm_corr == "LaSRC")) %>%
  left_join(., am_corr_LS7) %>%
  mutate(red_corr7 = med_Red_intercept + med_Red_B1*med_Red + med_Red_B2*med_Red^2,
         green_corr7 = med_Green_intercept + med_Green_B1*med_Green + med_Green_B2*med_Green^2,
         blue_corr7 = med_Blue_intercept + med_Blue_B1*med_Blue + med_Blue_B2*med_Blue^2,
         nir_corr7 = med_Nir_intercept + med_Nir_B1*med_Nir + med_Nir_B2*med_Nir^2,
         swir1_corr7 = med_Swir1_intercept + med_Swir1_B1*med_Swir1 + med_Swir1_B2*med_Swir1^2,
         swir2_corr7 = med_Swir2_intercept + med_Swir2_B1*med_Swir2 + med_Swir2_B2*med_Swir2^2) %>% 
  # store ls7 in the corr7 column
  mutate(red_corr7 = if_else(sat_corr == "LS7", med_Red, red_corr7),
         green_corr7 = if_else(sat_corr == "LS7", med_Green, green_corr7),
         blue_corr7 = if_else(sat_corr == "LS7", med_Blue, blue_corr7),
         nir_corr7 = if_else(sat_corr == "LS7", med_Nir, nir_corr7),
         swir1_corr7 = if_else(sat_corr == "LS7", med_Swir1, swir1_corr7),
         swir2_corr7 = if_else(sat_corr == "LS7", med_Swir2, swir2_corr7)) %>% 
  select(-c(all_of(names(am_corr_LS7)))) 

```

Do a quick plot to check that the correction seems right

```{r}
max_n_sites <- split_dataframes_corr %>% 
  summarize(n = n(), .by = siteSR_id) %>% 
  filter(n > 100) %>% 
  left_join(split_dataframes_corr)

ggplot() +
  geom_point(data = max_n_sites, aes(x = date, y = med_Blue), color = "grey") +
  geom_point(data = max_n_sites, aes(x = date, y = blue_corr7, color = mission))  +
  facet_wrap(siteSR_id ~ .) +
  theme_bw()
```

## RANSAC-like filtering on the reflectance bands

Script 01 runs a RANSAC-like filter on SDD alone (one scalar, one line-vs-time
fit, one residual per point). The corrected bands here are six correlated
values per observation, so filtering them independently, band by band, would
let each band pick its own inlier subset - a single matchup row needs one
keep/drop decision, not six that disagree. Reflectance is also a shakier fit
against a straight line over a site's full record than SDD is (blooms,
turbidity, and seasonality move bands around much more), so instead of a raw
residual threshold in reflectance units (which wouldn't transfer across bands
on different scales anyway) we standardize each band's per-iteration residual
by its own MAD and combine into one z-score per point, then threshold on that
combined score. Same random-subsample-and-count-inliers structure as
`ransac_site()`, just fitting all six bands per iteration instead of one.

Critically, each band's z-score scale is estimated *once*, from a single
all-data fit, rather than re-derived from each candidate subsample's own
residuals: standardizing by a per-iteration scale lets a bad candidate line
inflate its own denominator, so a wild-enough fit makes every point
(including the real outliers) look small in z-units and wins on raw inlier
count. A fixed reference scale keeps a bad candidate's residuals large
relative to a stable yardstick, so it can't win that way.

```{r ransac-multiband-function}
ransac_site_multiband <- function(date, band_mat, n_iter = 200, threshold = 2.5,
                                  min_sample = 8) {
  n <- nrow(band_mat)
  nb <- ncol(band_mat)
  if (n < min_sample + 1) return(rep(FALSE, n))  # too few points to do anything

  t <- as.numeric(date)

  # fixed per-band residual scale from a single all-data fit (MAD is robust
  # up to ~50% contamination, well above what we expect here) - reused for
  # every candidate subsample below instead of recomputed per iteration
  band_scale <- vapply(seq_len(nb), function(b) {
    resid <- residuals(lm(band_mat[, b] ~ t))
    s <- mad(resid, na.rm = TRUE)
    if (is.na(s) || s == 0) s <- sd(resid, na.rm = TRUE)
    s
  }, numeric(1))

  best_inliers <- rep(FALSE, n)
  best_count <- 0

  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)

    # fit each band vs time on the *same* subsample, standardize each band's
    # residuals by its fixed scale, then combine - keeps one band's scale
    # (e.g. NIR) from dominating and keeps the threshold comparable across
    # candidate lines of very different quality
    z <- matrix(NA_real_, nrow = n, ncol = nb)
    for (b in seq_len(nb)) {
      y <- band_mat[, b]
      if (length(unique(y[idx])) < 2 || is.na(band_scale[b]) || band_scale[b] == 0) next

      fit <- lm(y[idx] ~ t[idx])
      pred <- coef(fit)[1] + coef(fit)[2] * t
      resid <- y - pred
      z[, b] <- abs(resid) / band_scale[b]
    }

    combined_z <- rowMeans(z, na.rm = TRUE)
    inliers <- !is.na(combined_z) & combined_z < threshold

    if (sum(inliers) > best_count) {
      best_count <- sum(inliers)
      best_inliers <- inliers
    }
  }
  best_inliers
}
```

Apply it per site (`siteSR_id`) on the six sensor-corrected bands, arranged by
date so the time fit is meaningful.

```{r ransac-multiband-run}
set.seed(47)

band_cols <- c("red_corr7", "green_corr7", "blue_corr7",
               "nir_corr7", "swir1_corr7", "swir2_corr7")

split_dataframes_corr <- split_dataframes_corr %>%
  arrange(siteSR_id, date) %>%
  group_by(siteSR_id) %>%
  mutate(is_band_inlier = ransac_site_multiband(date, as.matrix(pick(all_of(band_cols))))) %>%
  ungroup()

# how many observations (and what fraction) got flagged per site
split_dataframes_corr %>%
  summarise(n = n(), n_outlier = sum(!is_band_inlier), .by = siteSR_id) %>%
  filter(n_outlier > 0) %>%
  mutate(frac_outlier = n_outlier / n) %>%
  arrange(desc(frac_outlier))
```

Spot-check on the same high-record sites plotted above: flagged points should
look like genuine spectral excursions (residual cloud/shadow contamination,
glint) rather than a large fraction of any one site's record.

```{r ransac-multiband-check}
max_n_sites_flagged <- split_dataframes_corr %>%
  filter(siteSR_id %in% unique(max_n_sites$siteSR_id))

ggplot(max_n_sites_flagged, aes(x = date, y = nir_corr7, color = is_band_inlier)) +
  geom_point() +
  facet_wrap(siteSR_id ~ .) +
  scale_color_manual(values = c(`TRUE` = "grey40", `FALSE` = "red")) +
  theme_bw()
```

Drop the flagged rows before moving on to spectral-index engineering.

```{r ransac-multiband-filter}
split_dataframes_corr <- split_dataframes_corr %>%
  filter(is_band_inlier) %>%
  select(-is_band_inlier)
```

## Engineer spectral indices

Build band-ratio and normalized-difference features from the
sensor-corrected bands for use as model predictors.

```{r spectral-indices}
split_dataframes_ml <- split_dataframes_corr %>%
  mutate(NR = nir_corr7/red_corr7,
         BR = blue_corr7/red_corr7,
         GR = green_corr7/red_corr7,
         SR = swir1_corr7/red_corr7,
         BG = blue_corr7/green_corr7,
         RG = red_corr7/green_corr7,
         NG = nir_corr7/green_corr7,
         SG = swir1_corr7/green_corr7,
         BN = blue_corr7/nir_corr7,
         GN = green_corr7/nir_corr7,
         RN = red_corr7/nir_corr7,
         SN = swir1_corr7/nir_corr7,
         BS = blue_corr7/swir1_corr7,
         GS = green_corr7/swir1_corr7,
         RS = red_corr7/swir1_corr7,
         NS = nir_corr7/swir1_corr7,
         R_GN = red_corr7/(green_corr7 + nir_corr7),
         R_GB = red_corr7/(green_corr7 + blue_corr7),
         R_GS = red_corr7/(green_corr7 + swir1_corr7),
         R_BN = red_corr7/(blue_corr7 + nir_corr7),
         R_BS = red_corr7/(blue_corr7 + swir1_corr7),
         R_NS = red_corr7/(nir_corr7 + swir1_corr7),
         G_BR = green_corr7/(blue_corr7 + swir1_corr7),
         G_BN = green_corr7/(blue_corr7 + nir_corr7),
         G_BS = green_corr7/(blue_corr7 + swir1_corr7),
         G_RN = green_corr7/(red_corr7 + nir_corr7),
         G_RB = green_corr7/(red_corr7 + blue_corr7),
         G_NS = green_corr7/(nir_corr7 + swir1_corr7),
         B_RG = blue_corr7/(red_corr7 + green_corr7),
         B_RS = blue_corr7/(red_corr7 + swir1_corr7),
         B_GN = blue_corr7/(green_corr7 + nir_corr7),
         B_GS = blue_corr7/(green_corr7 + swir1_corr7),
         B_NS = blue_corr7/(nir_corr7 + swir1_corr7),
         N_RG = nir_corr7/(red_corr7 + green_corr7),
         N_RB = nir_corr7/(red_corr7 + blue_corr7),
         N_RS = nir_corr7/(red_corr7 + swir1_corr7),
         N_GB = nir_corr7/(green_corr7 + blue_corr7),
         N_GS = nir_corr7/(green_corr7 + nir_corr7),
         N_BS = nir_corr7/(blue_corr7 + swir1_corr7),
         GR_2 = (red_corr7 + green_corr7)/2,
         GN_2 = (nir_corr7 + green_corr7)/2,
         BR_G = (blue_corr7 - red_corr7)/green_corr7,
         NS_NR = (nir_corr7 - swir1_corr7)/(red_corr7 - swir1_corr7),
         fai = nir_corr7 - (red_corr7 + (swir1_corr7 - red_corr7) * ((830-660)/(1650-660))),
         NmS = nir_corr7 - swir1_corr7,
         NmR = nir_corr7 - red_corr7,
         NDVI = (nir_corr7 - red_corr7)/(nir_corr7 + red_corr7),
         NDWI = (green_corr7 - swir1_corr7)/(green_corr7 + swir1_corr7),
         NDSSI = (blue_corr7 - nir_corr7)/(blue_corr7 + nir_corr7),
         GN_GN = (green_corr7- nir_corr7)/(green_corr7 + nir_corr7)) %>%
  # a handful of corrected bands land on exactly 0 (or two bands land on the
  # same value), which sends some ratio/difference features to Inf/NaN;
  # xgb.DMatrix() rejects those outright, so recode as NA and let xgboost's
  # native missing-value handling take it from there
  mutate(across(red_corr7:GN_GN, ~ if_else(is.finite(.x), .x, NA_real_)))
```

## Create train/validation/test splits

We use a leave-one-partition-out cross-validation scheme across the four
partitions assigned in script 03 (`part` 1–4), holding partition 5 out
entirely as an untouched test set.

```{r train-val-splits}
train_val <- split_dataframes_ml %>%
  filter(part != 4)

train1 <- train_val %>%
  filter(part != 1)
val1 <- anti_join(train_val, train1)

train2 <- train_val %>%
  filter(part != 2)
val2 <- anti_join(train_val, train2)

train3 <- train_val %>%
  filter(part != 3)
val3 <- anti_join(train_val, train3)

train4<- train_val %>%
  filter(part != 5)
val4 <- anti_join(train_val, train4)
```

Assemble the corrected bands and spectral indices as model features and
build XGBoost DMatrices for each fold.

```{r dmatrices}
target <- "harmonized_value"

feats_candidate <-  train_val %>% select(red_corr7:GN_GN) %>% names(.) %>% c("atm_corr_LaSRC")

dtrain1 <- xgb.DMatrix(data = as.matrix(train1[,feats_candidate]),
                       label = train1$harmonized_value)

dval1 <- xgb.DMatrix(data = as.matrix(val1[,feats_candidate]),
                     label = val1$harmonized_value)

dtrain2 <- xgb.DMatrix(data = as.matrix(train2[,feats_candidate]),
                       label = train2$harmonized_value)

dval2 <- xgb.DMatrix(data = as.matrix(val2[,feats_candidate]),
                     label = val2$harmonized_value)

dtrain3 <- xgb.DMatrix(data = as.matrix(train3[,feats_candidate]),
                       label = train3$harmonized_value)

dval3 <- xgb.DMatrix(data = as.matrix(val3[,feats_candidate]),
                     label = val3$harmonized_value)

dtrain4 <- xgb.DMatrix(data = as.matrix(train4[,feats_candidate]),
                       label = train4$harmonized_value)

dval4 <- xgb.DMatrix(data = as.matrix(val4[,feats_candidate]),
                     label = val4$harmonized_value)
```

Build the held-out test split (partition 4).

```{r test-split}
test <- split_dataframes_ml %>%
  filter(part == 4)

dtest <- xgb.DMatrix(data = as.matrix(test[,feats_candidate]),
                     label = test$harmonized_value)
```

## Explore baseline model performance

Quick sanity-check fits with a tree booster and a linear booster, just to
get a feel for error magnitude and feature importance before tuning.

```{r naive-gbtree}
params <- list(booster = "gbtree",
               objective = 'reg:squarederror',
               eta=0.01,
               max_depth=7,
               min_child_weight=1,
               subsample=0.8,
               colsample_bytree=0.5)

xgb.naive <- xgb.train(params = params, data = dtrain1, nrounds = 2000,
                       evals = list(train = dtrain1, val = dval1),
                       verbose = 1,
                       early_stopping_rounds = 100)

importance <- xgb.importance(model = xgb.naive, feature_names = feats_candidate)
xgb.plot.importance(importance_matrix = importance)
```

```{r naive-gblinear}
params_linear <- list(booster = "gblinear",
                      objective = 'reg:squarederror',
                      eta=0.1,
                      max_depth=7,
                      min_child_weight=1,
                      subsample=0.8,
                      colsample_bytree=0.5)

xgb.naive.2 <- xgb.train(params = params_linear,
                         data = dtrain1,
                         nrounds = 5000,
                         evals = list(train = dtrain1, val = dval1),
                         verbose = 1,
                         early_stopping_rounds = 100)

importance.2 <- xgb.importance(model = xgb.naive.2, feature_names = feats_candidate)
xgb.plot.importance(importance_matrix = importance.2)
```

As expected, the linear tuner is much worse than the tree, so we'll just stick
with the tree version.

## Prune correlated features

Many of the `r length(feats_candidate)` candidate features are near-duplicate
recombinations of the same 6 corrected bands, so before we spend any time
tuning or eliminating features one at a time we collapse obviously-redundant
clusters. We compute pairwise correlations among all candidate features on
`train_val` only (the held-out test partition is never touched by any of this),
link any pair with `|r| > 0.95` into a graph, and take connected components as
clusters of mutually-redundant features. Within each cluster we keep the one
feature most correlated with `harmonized_value`; singletons pass through
untouched.

```{r corr-pruning}
corr_threshold <- 0.95

cor_mat <- train_val %>%
  select(all_of(feats_candidate)) %>%
  as.matrix() %>%
  cor(use = "pairwise.complete.obs")

high_cor_idx <- which(abs(cor_mat) > corr_threshold & upper.tri(cor_mat), arr.ind = TRUE)

cor_edges <- tibble(from = rownames(cor_mat)[high_cor_idx[, 1]],
                    to = colnames(cor_mat)[high_cor_idx[, 2]])

cor_graph <- graph_from_data_frame(cor_edges, directed = FALSE, vertices = feats_candidate)
cor_clusters <- components(cor_graph)$membership

target_cor <- map_dbl(feats_candidate, ~ abs(cor(train_val[[.x]], train_val$harmonized_value,
                                                 use = "pairwise.complete.obs")))

cluster_tbl <- tibble(feature = names(cor_clusters),
                      cluster = cor_clusters,
                      target_cor = target_cor[names(cor_clusters)]) %>%
  arrange(cluster, desc(target_cor))

feats_decorr <- cluster_tbl %>%
  slice_max(target_cor, n = 1, by = cluster, with_ties = FALSE) %>%
  pull(feature)

# audit: which features got dropped, and in favor of which cluster-mate
cluster_tbl %>%
  mutate(kept = feature %in% feats_decorr) %>%
  filter(n() > 1, .by = cluster) %>%
  arrange(cluster, desc(kept))
```

Correlation pruning took the candidate set from `r length(feats_candidate)`
down to `r length(feats_decorr)` features.

## Backward elimination with cross-fold consensus

We eliminate features one at a time using the same 4-fold leave-one-partition-out
CV structure built above (`train1`/`val1` … `train4`/`val4`, i.e. partitions
1, 2, 3, and 5 — the test partition is still never touched). At each step we
fit all 4 folds on the current feature set, decide which feature to drop next
by averaging its importance (`Gain`) *across* the 4 folds rather than trusting
any single fold, and record how many folds individually agree with that
consensus pick, so disagreement between splits is visible rather than
averaged away.

```{r elimination-function}
elim_params <- list(booster = "gbtree",
                    objective = 'reg:squarederror',
                    eta = 0.05,
                    max_depth = 4,
                    min_child_weight = 3,
                    subsample = 0.8,
                    colsample_bytree = 0.8)

fit_fold <- function(train_df, val_df, feats, params, nrounds = 800, early_stopping_rounds = 25) {
  dtr <- xgb.DMatrix(data = as.matrix(train_df[, feats]), label = train_df$harmonized_value)
  dv  <- xgb.DMatrix(data = as.matrix(val_df[, feats]), label = val_df$harmonized_value)
  
  fit <- xgb.train(params = params, data = dtr, nrounds = nrounds,
                   evals = list(train = dtr, val = dv),
                   verbose = 0,
                   early_stopping_rounds = early_stopping_rounds)
  
  list(rmse = unname(attr(fit, "early_stop")$best_score),
       importance = xgb.importance(model = fit, feature_names = feats))
}

elim_folds <- list(list(train = train1, val = val1),
                   list(train = train2, val = val2),
                   list(train = train3, val = val3),
                   list(train = train4, val = val4))
```

```{r elimination-run}
set.seed(47)

min_feats <- 3
current_feats <- feats_decorr
history <- list()
step <- 1

repeat {
  fold_fits <- map(elim_folds, ~ fit_fold(.x$train, .x$val, current_feats, elim_params))
  fold_rmse <- map_dbl(fold_fits, "rmse")
  
  imp_long <- map2_dfr(fold_fits, seq_along(fold_fits),
                       ~ mutate(.x$importance, fold = .y)) %>%
    select(Feature, fold, Gain) %>%
    complete(Feature = current_feats, fold = seq_along(elim_folds), fill = list(Gain = 0)) %>%
    mutate(rank = rank(Gain), .by = fold)
  
  consensus <- imp_long %>%
    summarise(mean_gain = mean(Gain), .by = Feature) %>%
    arrange(mean_gain)
  
  drop_feat <- consensus$Feature[1]
  concordance <- imp_long %>%
    filter(Feature == drop_feat, rank <= 3) %>%
    nrow()
  
  history[[step]] <- tibble(step = step,
                            n_features = length(current_feats),
                            mean_rmse = mean(fold_rmse),
                            se_rmse = sd(fold_rmse) / sqrt(length(fold_rmse)),
                            dropped_feature = drop_feat,
                            concordance = concordance,
                            fold_rmse = list(fold_rmse),
                            retained_features = list(current_feats),
                            fold_importance = list(imp_long))
  
  if (length(current_feats) <= min_feats) break
  
  current_feats <- setdiff(current_feats, drop_feat)
  step <- step + 1
}

elim_history <- bind_rows(history)
```

Plot the elimination curve (mean validation RMSE across the 4 folds, ± 1 SE,
as features are removed).

```{r elimination-plot}
ggplot(elim_history, aes(x = n_features, y = mean_rmse)) +
  geom_ribbon(aes(ymin = mean_rmse - se_rmse, ymax = mean_rmse + se_rmse), alpha = 0.2) +
  geom_line() +
  geom_point() +
  scale_x_reverse() +
  labs(x = "number of features retained",
       y = "mean CV RMSE (4 folds)") +
  theme_bw()
```

An automatic rule (e.g. 1-SE off the minimum) is too easily fooled here: with
only 4 folds, `se_rmse` is a noisy estimate and can be wide enough to make a
real, visible rise in RMSE look like statistical noise, pulling the "optimal"
count all the way down to `min_feats`. Instead, read the elimination curve
above and set `n_features_selected` to the point where mean CV RMSE starts
climbing (the visible knee), then check how much the 4 folds agree at that
step — both on overall performance (fold RMSE spread) and on which retained
features actually matter to each fold individually (Gain by fold).

```{r elimination-select}
# set this by eye from the elimination-plot above: the number of features
# right before mean CV RMSE starts climbing
n_features_selected <- 9

stopifnot("n_features_selected must match a step in elim_history" =
            n_features_selected %in% elim_history$n_features)

selected_row <- elim_history %>%
  filter(n_features == n_features_selected)

feats_selected <- selected_row$retained_features[[1]]

# persist for reuse by 05_split_seed_sensitivity.Rmd, so that script doesn't
# need to re-run (eyeballed) elimination to get the same feature set
write_rds(feats_selected, "regional_clarity/xg_models/prelim/feats_selected.rds")

# fold RMSE spread at the selected step
tibble(fold = seq_along(selected_row$fold_rmse[[1]]),
       rmse = selected_row$fold_rmse[[1]])

# per-fold Gain for each retained feature at the selected step
selected_row$fold_importance[[1]] %>%
  select(Feature, fold, Gain) %>%
  pivot_wider(names_from = fold, names_prefix = "fold", values_from = Gain) %>%
  arrange(desc(rowSums(across(starts_with("fold")))))
```

Backward elimination selected `r length(feats_selected)` features (from
`r length(feats_decorr)` after correlation pruning, `r length(feats_candidate)`
originally).

Rebuild the fold DMatrices and the held-out test DMatrix using the selected
feature set — everything downstream (hyperparameter tuning, final training,
test evaluation) now runs on `feats_selected`.

```{r rebuild-dmatrices}
feats <- feats_selected

dtrain1 <- xgb.DMatrix(data = as.matrix(train1[,feats]),
                       label = train1$harmonized_value)
dval1 <- xgb.DMatrix(data = as.matrix(val1[,feats]),
                     label = val1$harmonized_value)

dtrain2 <- xgb.DMatrix(data = as.matrix(train2[,feats]),
                       label = train2$harmonized_value)
dval2 <- xgb.DMatrix(data = as.matrix(val2[,feats]),
                     label = val2$harmonized_value)

dtrain3 <- xgb.DMatrix(data = as.matrix(train3[,feats]),
                       label = train3$harmonized_value)
dval3 <- xgb.DMatrix(data = as.matrix(val3[,feats]),
                     label = val3$harmonized_value)

dtrain4 <- xgb.DMatrix(data = as.matrix(train4[,feats]),
                       label = train4$harmonized_value)
dval4 <- xgb.DMatrix(data = as.matrix(val4[,feats]),
                     label = val4$harmonized_value)

dtest <- xgb.DMatrix(data = as.matrix(test[,feats]),
                     label = test$harmonized_value)
```

## Hyperparameter tuning

We grid search `max_depth` and `min_child_weight` for the `gbtree` booster in
two rounds. Each grid point is fit across all 4 CV folds (not just fold 1) so
the score reflects how well a parameter combination generalizes across
partitions rather than how well it happens to fit one — same cross-fold
logic as the feature elimination step above, and still only ever touching
`dtrain1..4`/`dval1..4` (partitions 1, 2, 3, 5), never the held-out test
partition. Alongside `mean_val_loss`, each grid point also reports
`mean_train_loss` and `mean_gap` (mean val − train RMSE across folds, at the
same early-stopped iteration used for `mean_val_loss`) — useful for spotting
combinations that look good on validation loss alone but are already
overfitting relative to their own training data.


```{r hypertune-function}
tune_folds <- list(list(train = dtrain1, val = dval1),
                   list(train = dtrain2, val = dval2),
                   list(train = dtrain3, val = dval3),
                   list(train = dtrain4, val = dval4))

hypertune_xgboost = function(folds, grid){
  
  params <- list(booster = "gbtree",
                 objective = 'reg:squarederror',
                 eta=grid$eta,
                 max_depth=grid$max_depth,
                 min_child_weight=grid$min_child_weight,
                 subsample=grid$subsample,
                 colsample_bytree=grid$colsample_bytree)
  
  fold_results <- map(folds, function(fold) {
    xgb <- xgb.train(params = params, data = fold$train, nrounds = 5000,
                     evals = list(train = fold$train, val = fold$val),
                     verbose = 0,
                     early_stopping_rounds = 100)
    
    # as of xgboost >=2.0, fit results (best_score, best_iteration) live in
    # attributes rather than list elements, so xgb$best_score no longer works
    early_stop <- attr(xgb, "early_stop")
    
    # train RMSE at that same (early-stopped) iteration, so the train/val gap
    # reflects the model actually being scored rather than wherever training
    # happened to stop logging; as of xgboost >=2.0, evaluation_log also
    # lives in an attribute rather than a list element
    train_loss <- attr(xgb, "evaluation_log") %>%
      filter(iter == early_stop$best_iteration) %>%
      pull(train_rmse)
    
    list(val_loss = unname(early_stop$best_score),
         train_loss = train_loss,
         best_iteration = early_stop$best_iteration)
  })
  
  val_losses <- map_dbl(fold_results, "val_loss")
  train_losses <- map_dbl(fold_results, "train_loss")
  gaps <- val_losses - train_losses
  
  summary <- grid %>% mutate(mean_val_loss = mean(val_losses),
                             se_val_loss = sd(val_losses) / sqrt(length(val_losses)),
                             mean_train_loss = mean(train_losses),
                             # val - train: how much worse a combo does on
                             # held-out data than on the data it was fit to;
                             # a bigger gap is a sign of overfitting even if
                             # mean_val_loss itself looks competitive
                             mean_gap = mean(gaps),
                             se_gap = sd(gaps) / sqrt(length(gaps)),
                             fold_val_loss = list(val_losses),
                             fold_train_loss = list(train_losses),
                             mean_best_iteration = mean(map_dbl(fold_results, "best_iteration")))
  
  return(summary)
  
}
```

Round 1: a broad grid over `max_depth`, `min_child_weight`, `eta`, and
`colsample_bytree`. `max_depth`'s upper range is narrower than a first pass
would use on the full ~50-feature candidate set, since with only a handful of
features surviving pruning/elimination, deep trees (7+) mostly just cost
runtime rather than add anything. `eta` and `colsample_bytree` are searched
here rather than fixed, since the final training run uses a much lower `eta`
than earlier drafts assumed, and `colsample_bytree` was tuned back when there
were far more (collinear) candidate features than there are now — both are
worth re-checking against the current, pruned feature set rather than
carrying old defaults forward. This grid is 5 × 4 × 3 × 3 = 180 combinations
× 4 folds = 720 fits.

```{r hypertune-round1}
param_test1 <- expand.grid(
  booster = "gbtree",
  objective = 'reg:squarederror',
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(2, 3, 4, 5, 6),
  min_child_weight = c(1, 3, 5, 7),
  subsample = 0.8,
  colsample_bytree = c(0.5, 0.7, 1)
) %>%
  rowid_to_column()

## Hypertune xgboost for max depth, child weight, eta, and colsample_bytree
xgboost_hypertune <- param_test1 %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(tune_folds, current)
  })

xgboost_hypertune %>%
  arrange(mean_val_loss) %>%
  select(-c(fold_val_loss, fold_train_loss))
```

Round 2: refine around round 1's best combination. `eta` is locked to round
1's winning value (its 3-point grid was only meant to find the right
ballpark, not to be refined further), while `max_depth`, `min_child_weight`,
and `colsample_bytree` get a tight neighborhood search — built
programmatically off round 1's actual winner rather than re-typed by hand, so
this stays correct if the upstream feature set (and therefore round 1's
result) changes on a rerun.

```{r hypertune-round2}
round1_best <- xgboost_hypertune %>%
  slice_min(mean_val_loss, n = 1, with_ties = FALSE)

param_test2 <- expand.grid(
  booster = "gbtree",
  objective = 'reg:squarederror',
  eta = round1_best$eta,
  max_depth = unique(pmax(2, round1_best$max_depth + c(-1, 0, 1))),
  min_child_weight = unique(pmax(1, round1_best$min_child_weight + c(-2, 0, 2))),
  subsample = 0.8,
  colsample_bytree = unique(pmin(1, pmax(0.3, round1_best$colsample_bytree + c(-0.15, 0, 0.15))))
) %>%
  rowid_to_column()

xgboost_hypertune2 <- param_test2 %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(tune_folds, current)
  })

xgboost_hypertune2 %>%
  arrange(mean_val_loss) %>%
  select(-c(fold_val_loss, fold_train_loss))
```

## Train and cross-validate final models

We take round 2's winning combination directly rather than hand-copying
numbers in, so this stays correct if the tuning grid or upstream feature set
changes on a rerun. The pick below is just the lowest `mean_val_loss`;
`mean_gap` (mean val − train RMSE across folds) is included alongside it so
you can sanity-check overfitting risk before trusting that pick — if a
combination just behind it on `mean_val_loss` has a noticeably smaller
`mean_gap`, override `final_hyperparams` below with that row's `rowid`
instead.

```{r select-final-hyperparams}
final_hyperparams <- xgboost_hypertune2 %>%
  slice_min(mean_val_loss, n = 1, with_ties = FALSE)

# to override with a different row after weighing mean_gap against
# mean_val_loss in the table above, e.g.:
# final_hyperparams <- xgboost_hypertune2 %>% filter(rowid == <id>)

# persist for reuse by 05_split_seed_sensitivity.Rmd, so that script doesn't
# need to re-run hyperparameter tuning to get the same configuration
write_rds(final_hyperparams, "regional_clarity/xg_models/prelim/final_hyperparams.rds")

final_hyperparams %>%
  select(-c(fold_val_loss, fold_train_loss))
```

```{r train-final-models}
params <- list(booster = "gbtree",
               objective = 'reg:squarederror',
               eta = final_hyperparams$eta,
               max_depth = final_hyperparams$max_depth,
               min_child_weight = final_hyperparams$min_child_weight,
               subsample = final_hyperparams$subsample,
               colsample_bytree = final_hyperparams$colsample_bytree)

set.seed(47)
xgb1 <- xgb.train(params = params, data = dtrain1, nrounds = 5000,
                  evals = list(train = dtrain1, val = dval1),
                  verbose = 1,
                  early_stopping_rounds = 250)
xgb.save(xgb1, "regional_clarity/xg_models/prelim/xgb1.model")
xgb2 <- xgb.train(params = params, data = dtrain2, nrounds = 5000,
                  evals = list(train = dtrain2, val = dval2),
                  verbose = 1,
                  early_stopping_rounds = 250)
xgb.save(xgb2, "regional_clarity/xg_models/prelim/xgb2.model")
xgb3 <- xgb.train(params = params, data = dtrain3, nrounds = 5000,
                  evals = list(train = dtrain3, val = dval3),
                  verbose = 1,
                  early_stopping_rounds = 250)
xgb.save(xgb3, "regional_clarity/xg_models/prelim/xgb3.model")
xgb4 <- xgb.train(params = params, data = dtrain4, nrounds = 5000,
                  evals = list(train = dtrain4, val = dval4),
                  verbose = 1,
                  early_stopping_rounds = 250)
xgb.save(xgb4, "regional_clarity/xg_models/prelim/xgb4.model")
```

For each fold, average the predictions from all four models and compare
against the fold's own held-out validation set.

```{r validate-fold1}
pred_1 <- val1 %>%
  mutate(pred1 = predict(xgb1, dval1),
         pred2 = predict(xgb2, dval1),
         pred3 = predict(xgb3, dval1),
         pred4 = predict(xgb4, dval1)) %>%
  rowwise() %>%
  mutate(mean = sum(pred1, pred2, pred3, pred4)/4)

ggplot(pred_1, aes(x = harmonized_value, y = mean)) +
  geom_point() +
  geom_point(aes(y = pred1), color = "yellow") +
  theme_bw()
```

```{r validate-fold2}
pred_2 <- val2 %>%
  mutate(pred1 = predict(xgb1, dval2),
         pred2 = predict(xgb2, dval2),
         pred3 = predict(xgb3, dval2),
         pred4 = predict(xgb4, dval2)) %>%
  rowwise() %>%
  mutate(mean = sum(pred1, pred2, pred3, pred4)/4)

ggplot(pred_2, aes(x = harmonized_value, y = mean)) +
  geom_point() +
  geom_point(aes(y = pred2), color = "yellow") +
  theme_bw()
```

```{r validate-fold3}
pred_3 <- val3 %>%
  mutate(pred1 = predict(xgb1, dval3),
         pred2 = predict(xgb2, dval3),
         pred3 = predict(xgb3, dval3),
         pred4 = predict(xgb4, dval3)) %>%
  rowwise() %>%
  mutate(mean = sum(pred1, pred2, pred3, pred4)/4)

ggplot(pred_3, aes(x = harmonized_value, y = mean)) +
  geom_point() +
  geom_point(aes(y = pred3), color = "yellow") +
  theme_bw()
```

```{r validate-fold4}
pred_4 <- val4 %>%
  mutate(pred1 = predict(xgb1, dval4),
         pred2 = predict(xgb2, dval4),
         pred3 = predict(xgb3, dval4),
         pred4 = predict(xgb4, dval4)) %>%
  rowwise() %>%
  mutate(mean = sum(pred1, pred2, pred3, pred4)/4)

ggplot(pred_4, aes(x = harmonized_value, y = mean)) +
  geom_point() +
  geom_point(aes(y = pred4), color = "yellow") +
  theme_bw()
```

## Re-weight training toward high-SDD values

Across all four folds above, predictions increasingly fall below the 1:1
line as observed SDD grows — a "ceiling" effect that's common when a
squared-error loss is fit to a right-skewed target: the dense cluster of
low/mid-SDD observations dominates the loss, so the model is never pushed
hard to get the sparse high-SDD tail right. Here we refit each fold with
per-observation training weights that scale linearly with SDD, so a
high-SDD point counts for more during training, likely at a small cost to
overall (low/mid-range) accuracy. This is a comparison, not a replacement:
the unweighted `xgb1`-`xgb4` above are left untouched and still carry
through to the held-out test evaluation below; the weighted models here are
a parallel set (`xgb1_w`-`xgb4_w`).

```{r sdd-weight-function}
# weight strength: at weight_strength = k, the highest-SDD training
# observation in a fold is weighted (1 + k) times as heavily as the lowest;
# tune by eye against the comparison metrics/plot below - start modest
# since this trades off overall accuracy for high-SDD accuracy
weight_strength <- 2

sdd_weight <- function(value, k = weight_strength) {
  rng <- range(value, na.rm = TRUE)
  1 + k * (value - rng[1]) / (rng[2] - rng[1])
}
```

Build training DMatrices with these weights attached. Validation DMatrices
stay the existing unweighted `dval1..4`, so early stopping and the reported
val-rmse remain directly comparable to the unweighted models above.

```{r weighted-dmatrices}
dtrain1_w <- xgb.DMatrix(data = as.matrix(train1[,feats]),
                         label = train1$harmonized_value,
                         weight = sdd_weight(train1$harmonized_value))
dtrain2_w <- xgb.DMatrix(data = as.matrix(train2[,feats]),
                         label = train2$harmonized_value,
                         weight = sdd_weight(train2$harmonized_value))
dtrain3_w <- xgb.DMatrix(data = as.matrix(train3[,feats]),
                         label = train3$harmonized_value,
                         weight = sdd_weight(train3$harmonized_value))
dtrain4_w <- xgb.DMatrix(data = as.matrix(train4[,feats]),
                         label = train4$harmonized_value,
                         weight = sdd_weight(train4$harmonized_value))
```

Retrain each fold using the same tuned hyperparameters and early-stopping
settings as the unweighted final models above, changing only the training
data to add weights.

```{r train-weighted-models}
xgb1_w <- xgb.train(params = params, data = dtrain1_w, nrounds = 5000,
                    evals = list(train = dtrain1_w, val = dval1),
                    verbose = 1,
                    early_stopping_rounds = 250)
xgb.save(xgb1_w, "regional_clarity/xg_models/prelim/xgb1_weighted.model")
xgb2_w <- xgb.train(params = params, data = dtrain2_w, nrounds = 5000,
                    evals = list(train = dtrain2_w, val = dval2),
                    verbose = 1,
                    early_stopping_rounds = 250)
xgb.save(xgb2_w, "regional_clarity/xg_models/prelim/xgb2_weighted.model")
xgb3_w <- xgb.train(params = params, data = dtrain3_w, nrounds = 5000,
                    evals = list(train = dtrain3_w, val = dval3),
                    verbose = 1,
                    early_stopping_rounds = 250)
xgb.save(xgb3_w, "regional_clarity/xg_models/prelim/xgb3_weighted.model")
xgb4_w <- xgb.train(params = params, data = dtrain4_w, nrounds = 5000,
                    evals = list(train = dtrain4_w, val = dval4),
                    verbose = 1,
                    early_stopping_rounds = 250)
xgb.save(xgb4_w, "regional_clarity/xg_models/prelim/xgb4_weighted.model")
```

Add the weighted ensemble's predictions onto the existing fold prediction
tibbles, so both models' predictions live on the same rows for direct
comparison.

```{r weighted-fold-predictions}
pred_1 <- pred_1 %>%
  ungroup() %>%
  mutate(w_pred1 = predict(xgb1_w, dval1),
         w_pred2 = predict(xgb2_w, dval1),
         w_pred3 = predict(xgb3_w, dval1),
         w_pred4 = predict(xgb4_w, dval1),
         mean_w = (w_pred1 + w_pred2 + w_pred3 + w_pred4)/4)

pred_2 <- pred_2 %>%
  ungroup() %>%
  mutate(w_pred1 = predict(xgb1_w, dval2),
         w_pred2 = predict(xgb2_w, dval2),
         w_pred3 = predict(xgb3_w, dval2),
         w_pred4 = predict(xgb4_w, dval2),
         mean_w = (w_pred1 + w_pred2 + w_pred3 + w_pred4)/4)

pred_3 <- pred_3 %>%
  ungroup() %>%
  mutate(w_pred1 = predict(xgb1_w, dval3),
         w_pred2 = predict(xgb2_w, dval3),
         w_pred3 = predict(xgb3_w, dval3),
         w_pred4 = predict(xgb4_w, dval3),
         mean_w = (w_pred1 + w_pred2 + w_pred3 + w_pred4)/4)

pred_4 <- pred_4 %>%
  ungroup() %>%
  mutate(w_pred1 = predict(xgb1_w, dval4),
         w_pred2 = predict(xgb2_w, dval4),
         w_pred3 = predict(xgb3_w, dval4),
         w_pred4 = predict(xgb4_w, dval4),
         mean_w = (w_pred1 + w_pred2 + w_pred3 + w_pred4)/4)
```

Compare unweighted vs. weighted RMSE, split into the top quartile of SDD
(by `train_val`'s own distribution, never touching the test partition) vs.
the rest, to see whether the trade actually paid off and how much it cost.

```{r compare-weighted-metrics}
top_quartile_cut <- quantile(train_val$harmonized_value, 0.75, na.rm = TRUE)

compare_tbl <- bind_rows(pred_1 %>% mutate(fold = 1),
                         pred_2 %>% mutate(fold = 2),
                         pred_3 %>% mutate(fold = 3),
                         pred_4 %>% mutate(fold = 4)) %>%
  ungroup() %>%
  mutate(sdd_group = if_else(harmonized_value >= top_quartile_cut,
                             "top quartile", "rest"))

compare_tbl %>%
  summarise(rmse_unweighted = rmse(harmonized_value, mean),
            rmse_weighted = rmse(harmonized_value, mean_w),
            n = n(),
            .by = sdd_group)
```

Plot predicted vs. observed SDD for both models across all four validation
folds, to visually check whether the high tail moved closer to the 1:1 line.

```{r compare-weighted-plot}
compare_tbl %>%
  pivot_longer(c(mean, mean_w), names_to = "model", values_to = "prediction") %>%
  mutate(model = if_else(model == "mean", "unweighted", "SDD-weighted")) %>%
  ggplot(aes(x = harmonized_value, y = prediction, color = model)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, lty = 2, col = "grey40") +
  labs(x = "observed Secchi (validation folds)",
       y = "predicted Secchi",
       color = NULL) +
  theme_bw()
```

## Apply to held-out test partition

Predict on the fully held-out test partition (part 5) with each of the
four fold models and average, dropping any test observations flagged by
`misc_flag`.

```{r test-predictions}
test_pred <- test %>%
  mutate(pred1 = predict(xgb1_w, dtest),
         pred2 = predict(xgb2_w, dtest),
         pred3 = predict(xgb3_w, dtest),
         pred4 = predict(xgb4_w, dtest)) %>%
  rowwise() %>%
  mutate(mean = sum(pred1, pred2, pred3, pred4)/4) %>%
  filter(misc_flag != 1)

test_vert <- test_pred %>%
  pivot_longer(pred1:pred4,
               names_to = "pred",
               values_to = "values")

ggplot(test_vert, aes(x = harmonized_value, y = values)) +
  geom_point(aes(color = pred)) +
  scale_color_viridis_d() +
  labs(x = "observed Secchi",
       y = "predicted Secchi",
       color = "split") +
  theme_bw()

ggplot(test_vert, aes(x = harmonized_value, y = mean)) +
  geom_point() +
  labs(x = "observed Secchi",
       y = "predicted Secchi") +
  geom_abline(slope = 1, intercept = 0, lty = 1, col = "grey") +
  theme_bw()
```

Summarize test-set error with standard regression metrics.

```{r test-metrics}
evals_test <- test_pred %>%
  ungroup() %>%
  summarise(rmse = rmse(harmonized_value, mean),
            mae = mae(harmonized_value, mean),
            mape = mape(harmonized_value, mean),
            bias = bias(harmonized_value, mean),
            p.bias = percent_bias(harmonized_value, mean),
            smape = smape(harmonized_value, mean),
            r2 = cor(harmonized_value, mean)^2)
evals_test
```

Look at some longer test-set records over time, and their residuals, as a
sanity check.

```{r test-timeseries}
test_long <- test_pred %>%
  ungroup() %>%
  summarize(n = n(), .by = siteSR_id) %>%
  filter(n > 50) %>%
  left_join(., test_pred)

ggplot(test_long, aes(x = date, y = mean)) +
  geom_point() +
  geom_point(aes(y = harmonized_value), color = "light blue") +
  facet_grid(siteSR_id ~ ., scales = "free_y") +
  theme_bw()

ggplot(test_long, aes(x = date, y = mean - harmonized_value, color = HUC4)) +
  geom_point() +
  labs(y = "residual (m)",
       x = NULL) +
  theme_bw() +
  scale_color_viridis_d() +
  geom_abline(slope = 0, intercept = 0)
```

## Performance by HUC4

The residual plot above suggests error isn't uniform across HUC4 basins.
Break out the same test-set metrics computed above (`evals_test`), by HUC4,
across all held-out test observations (not just the longer-record subset
used for the time-series plot above), to see which basins the model
generalizes to well and which it doesn't.

```{r huc4-metrics}
evals_test_huc4 <- test_pred %>%
  ungroup() %>%
  summarise(n = n(),
            rmse = rmse(harmonized_value, mean),
            mae = mae(harmonized_value, mean),
            mape = mape(harmonized_value, mean),
            bias = bias(harmonized_value, mean),
            p.bias = percent_bias(harmonized_value, mean),
            smape = smape(harmonized_value, mean),
            r2 = cor(harmonized_value, mean)^2,
            .by = HUC4) %>%
  arrange(desc(rmse))

evals_test_huc4
```

Plot RMSE and bias by HUC4, ordered by RMSE, to see where error concentrates
and whether it's a systematic over/under-prediction or just a noisier fit.

```{r huc4-metrics-plot}
huc4_order <- evals_test_huc4 %>%
  arrange(rmse) %>%
  pull(HUC4)

evals_test_huc4 %>%
  mutate(HUC4 = factor(HUC4, levels = huc4_order)) %>%
  pivot_longer(c(rmse, bias), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = HUC4, y = value)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "grey40") +
  coord_flip() +
  facet_wrap(~metric, scales = "free_x") +
  labs(x = "HUC4", y = NULL) +
  theme_bw()
```

Map the same per-HUC4 metrics spatially so it's clear *where* the model
does and doesn't generalize well. HUC4 boundaries and the 6-state AOI are
built the same way script 03 did: HUC4's read from the national WBD
geodatabase (downloaded once and cached in `aquamatch_files/` by script 02),
filtered down to the basins that touch our region.

```{r huc4-boundaries}
states <- states(progress_bar = FALSE) %>%
  filter(STUSPS %in% c("CO", "WY", "UT", "ID", "MT", "NM"))
states_union <- states %>% st_union()

wbd_gdb <- download_wbd(outdir = "aquamatch_files/wbd_national")

huc4_layer <- st_layers(wbd_gdb)$name %>%
  keep(~ grepl("HU4$", ., ignore.case = TRUE))

HUC4s <- st_read(wbd_gdb, layer = huc4_layer, quiet = TRUE) %>%
  rename_with(tolower) %>%
  # WBD ships with Z values and a compound CRS (NAD83 + NAVD88 height);
  # transforming a 2D CRS into that compound CRS silently breaks (no vertical
  # grid to invent a Z), so drop Z here - horizontal datum is then plain NAD83
  st_zm(drop = TRUE, what = "ZM")

# standardize the geometry column name to "geometry" (FileGDB layers are
# often read in with the geometry column named "shape")
geom_col <- attr(HUC4s, "sf_column")
if (geom_col != "geometry") {
  names(HUC4s)[names(HUC4s) == geom_col] <- "geometry"
  st_geometry(HUC4s) <- "geometry"
}

# reset (not transform) the CRS now that only the horizontal NAD83 datum
# remains, matching states_union - no reprojection needed
st_crs(HUC4s) <- st_crs(states_union)

# st_make_valid() is by far the slowest step here (a handful of the WBD's
# 245 national HU4 polygons are topologically invalid, and validating fixes
# them all), so cheaply narrow to candidates near our AOI *before* validating
# rather than after. A spatial-index filter (st_filter(), `[`) would do that,
# but it invokes the same topology engine as st_make_valid() and errors on
# this dataset's invalid geometries - so instead compare each feature's own
# bounding box (plain coordinate min/max, no topology needed) against the
# AOI's bounding box.
aoi_bbox <- st_bbox(states_union)
feat_bboxes <- map(st_geometry(HUC4s), st_bbox)
near_aoi <- map_lgl(feat_bboxes, ~ !(.x["xmax"] < aoi_bbox["xmin"] ||
                                     .x["xmin"] > aoi_bbox["xmax"] ||
                                     .x["ymax"] < aoi_bbox["ymin"] ||
                                     .x["ymin"] > aoi_bbox["ymax"]))

# validate only those candidates, then do the precise (topological) filter
# down to the HUC4's that actually touch our 6-state AOI, and attach the
# test-set metrics by HUC4 code
huc4_map <- HUC4s[near_aoi, ] %>%
  st_make_valid() %>%
  st_filter(states_union) %>%
  select(huc4, geometry) %>%
  left_join(evals_test_huc4, by = c("huc4" = "HUC4"))
```

```{r huc4-map-rmse}
tm_shape(huc4_map) +
  tm_shape(states) +
  tm_polygons(col = "grey70") +
  tm_borders(col = "grey20") +
  tm_shape(huc4_map) +
  tm_polygons(fill = "rmse",
              fill.scale = tm_scale_continuous(values = "viridis"),
              fill.legend = tm_legend(title = "test RMSE (m)")) +
  tm_borders(col = "black") +
  tm_shape(states) +
  tm_text("STUSPS", col = "grey20") +
  tm_layout(frame = FALSE,
            main.title = "test-set RMSE by HUC4")
```

```{r huc4-map-bias}
tm_shape(huc4_map) +
  tm_shape(states) +
  tm_polygons(col = "grey70") +
  tm_borders(col = "grey20") +
  tm_shape(huc4_map) +
  tm_polygons(fill = "bias",
              fill.scale = tm_scale_continuous(midpoint = 0, values = "-RdBu"),
              fill.legend = tm_legend(title = "test bias (m)\n(+ = overprediction)")) +
  tm_borders(col = "black") +
  tm_shape(states) +
  tm_text("STUSPS", col = "grey20") +
  tm_layout(frame = FALSE,
            main.title = "test-set bias by HUC4")
```
