---
title: "Split & Seed Sensitivity"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 3, 2026"
output: html_document
---

Reviewers often ask how much reported test performance depends on (1)
which of script 03's 5 HUC4-grouped partitions happens to be held out as the
test set, and (2) the arbitrary choice of random seed. This script answers
both, reusing the feature set and hyperparameters already selected in
`04_make_models.Rmd` (read from disk below) rather than re-running the
eyeballed elimination/tuning steps for every scenario.

Because the feature set and hyperparameters are fixed rather than re-tuned
per held-out partition, the split-repetition results below characterize how
sensitive a *fixed, already-selected* modeling pipeline is to which data was
held out - not the sensitivity of the full model-selection process itself
(that would require re-running elimination and hyperparameter tuning inside
each split, which is both far more expensive and, as currently written,
requires eyeballing a plot at each repetition). This is the standard,
tractable compromise when full nested cross-validation is impractical, and
should be described as such in any reviewer response.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message = FALSE, warning=FALSE}
library(tidyverse)
library(arrow)
library(xgboost)
library(Metrics)

# read in the partitioned dataset from script 03
split_dataframes <- read_feather("aquamatch_files/split_filtered_regional.feather")

# feature set and hyperparameters selected once in 04_make_models.Rmd
feats <- read_rds("regional_clarity/xg_models/prelim/feats_selected.rds")
final_hyperparams <- read_rds("regional_clarity/xg_models/prelim/final_hyperparams.rds")
```

## Apply cross-sensor corrections and engineer spectral indices

Identical to the corresponding chunks in `04_make_models.Rmd`.

```{r sensor-corrections}
am_corr_LS7 <- read_csv("aquamatch_files/lakeSR_collated_handoffs_GEEv2025-02-12_QAv2025-06-04.csv") %>%
  filter(correction == "Gardner", dswe == "DSWE1", sat_to == "LS7") %>%
  pivot_longer(cols = c(intercept, B1, B2, min_in_val, max_in_val),
               names_to = "int_coef",
               values_to = "value") %>%
  mutate(new_column = paste(band, int_coef, sep = "_")) %>%
  select(-c(band, int_coef, slope, method)) %>%
  pivot_wider(names_from = new_column,
              values_from = value)

split_dataframes_corr <- split_dataframes %>%
  mutate(sat_corr = case_when(mission == "LT04" ~ "LS5",
                              mission == "LT05" ~ "LS5",
                              mission == "LE07" ~ "LS7",
                              mission == "LC08" ~ "LS8",
                              mission == "LC09" ~ "LS8",
                              TRUE ~ NA_character_),
         atm_corr = case_when(mission %in% c("LT04", "LT05", "LE07") ~ "LEDAPS",
                              mission %in% c("LC08", "LC09") ~ "LaSRC",
                              TRUE ~ NA_character_),
         atm_corr_LaSRC = as.numeric(atm_corr == "LaSRC")) %>%
  left_join(., am_corr_LS7) %>%
  mutate(red_corr7 = med_Red_intercept + med_Red_B1*med_Red + med_Red_B2*med_Red^2,
         green_corr7 = med_Green_intercept + med_Green_B1*med_Green + med_Green_B2*med_Green^2,
         blue_corr7 = med_Blue_intercept + med_Blue_B1*med_Blue + med_Blue_B2*med_Blue^2,
         nir_corr7 = med_Nir_intercept + med_Nir_B1*med_Nir + med_Nir_B2*med_Nir^2,
         swir1_corr7 = med_Swir1_intercept + med_Swir1_B1*med_Swir1 + med_Swir1_B2*med_Swir1^2,
         swir2_corr7 = med_Swir2_intercept + med_Swir2_B1*med_Swir2 + med_Swir2_B2*med_Swir2^2) %>%
  mutate(red_corr7 = if_else(sat_corr == "LS7", med_Red, red_corr7),
         green_corr7 = if_else(sat_corr == "LS7", med_Green, green_corr7),
         blue_corr7 = if_else(sat_corr == "LS7", med_Blue, blue_corr7),
         nir_corr7 = if_else(sat_corr == "LS7", med_Nir, nir_corr7),
         swir1_corr7 = if_else(sat_corr == "LS7", med_Swir1, swir1_corr7),
         swir2_corr7 = if_else(sat_corr == "LS7", med_Swir2, swir2_corr7)) %>%
  select(-c(all_of(names(am_corr_LS7))))
```


```{r spectral-indices}
split_dataframes_ml <- split_dataframes_corr %>%
  mutate(NR = nir_corr7/red_corr7,
         BR = blue_corr7/red_corr7,
         GR = green_corr7/red_corr7,
         SR = swir1_corr7/red_corr7,
         BG = blue_corr7/green_corr7,
         RG = red_corr7/green_corr7,
         NG = nir_corr7/green_corr7,
         SG = swir1_corr7/green_corr7,
         BN = blue_corr7/nir_corr7,
         GN = green_corr7/nir_corr7,
         RN = red_corr7/nir_corr7,
         SN = swir1_corr7/nir_corr7,
         BS = blue_corr7/swir1_corr7,
         GS = green_corr7/swir1_corr7,
         RS = red_corr7/swir1_corr7,
         NS = nir_corr7/swir1_corr7,
         R_GN = red_corr7/(green_corr7 + nir_corr7),
         R_GB = red_corr7/(green_corr7 + blue_corr7),
         R_GS = red_corr7/(green_corr7 + swir1_corr7),
         R_BN = red_corr7/(blue_corr7 + nir_corr7),
         R_BS = red_corr7/(blue_corr7 + swir1_corr7),
         R_NS = red_corr7/(nir_corr7 + swir1_corr7),
         G_BR = green_corr7/(blue_corr7 + swir1_corr7),
         G_BN = green_corr7/(blue_corr7 + nir_corr7),
         G_BS = green_corr7/(blue_corr7 + swir1_corr7),
         G_RN = green_corr7/(red_corr7 + nir_corr7),
         G_RB = green_corr7/(red_corr7 + blue_corr7),
         G_NS = green_corr7/(nir_corr7 + swir1_corr7),
         B_RG = blue_corr7/(red_corr7 + green_corr7),
         B_RS = blue_corr7/(red_corr7 + swir1_corr7),
         B_GN = blue_corr7/(green_corr7 + nir_corr7),
         B_GS = blue_corr7/(green_corr7 + swir1_corr7),
         B_NS = blue_corr7/(nir_corr7 + swir1_corr7),
         N_RG = nir_corr7/(red_corr7 + green_corr7),
         N_RB = nir_corr7/(red_corr7 + blue_corr7),
         N_RS = nir_corr7/(red_corr7 + swir1_corr7),
         N_GB = nir_corr7/(green_corr7 + blue_corr7),
         N_GS = nir_corr7/(green_corr7 + nir_corr7),
         N_BS = nir_corr7/(blue_corr7 + swir1_corr7),
         GR_2 = (red_corr7 + green_corr7)/2,
         GN_2 = (nir_corr7 + green_corr7)/2,
         BR_G = (blue_corr7 - red_corr7)/green_corr7,
         NS_NR = (nir_corr7 - swir1_corr7)/(red_corr7 - swir1_corr7),
         fai = nir_corr7 - (red_corr7 + (swir1_corr7 - red_corr7) * ((830-660)/(1650-660))),
         NmS = nir_corr7 - swir1_corr7,
         NmR = nir_corr7 - red_corr7,
         NDVI = (nir_corr7 - red_corr7)/(nir_corr7 + red_corr7),
         NDWI = (green_corr7 - swir1_corr7)/(green_corr7 + swir1_corr7),
         NDSSI = (blue_corr7 - nir_corr7)/(blue_corr7 + nir_corr7),
         GN_GN = (green_corr7- nir_corr7)/(green_corr7 + nir_corr7)) %>%
  mutate(across(red_corr7:GN_GN, ~ if_else(is.finite(.x), .x, NA_real_)))
```

## Repeated leave-one-partition-out training/evaluation

`run_split()` generalizes the leave-one-partition-out scheme from
`04_make_models.Rmd`: given which partition to hold out as the test set, it
builds the other 4 partitions into leave-one-out CV folds, trains one
XGBoost model per fold (same hyperparameters, early stopping, and rounds as
04's final training step), averages the 4 fold models' predictions on the
held-out test set, and returns test-set metrics. `test_partition` and `seed`
are both explicit arguments so the same function drives both experiments
below.

```{r run-split-function}
run_split <- function(data, feats, hyperparams, test_partition, seed,
                      nrounds = 5000, early_stopping_rounds = 250) {

  test <- data %>% filter(part == test_partition)
  train_val <- data %>% filter(part != test_partition)
  cv_partitions <- setdiff(sort(unique(data$part)), test_partition)

  params <- list(booster = "gbtree",
                 objective = "reg:squarederror",
                 eta = hyperparams$eta,
                 max_depth = hyperparams$max_depth,
                 min_child_weight = hyperparams$min_child_weight,
                 subsample = hyperparams$subsample,
                 colsample_bytree = hyperparams$colsample_bytree)

  dtest <- xgb.DMatrix(data = as.matrix(test[, feats]),
                       label = test$harmonized_value)

  set.seed(seed)
  fold_preds <- map(cv_partitions, function(p) {
    train_df <- train_val %>% filter(part != p)
    val_df   <- train_val %>% filter(part == p)

    dtr <- xgb.DMatrix(data = as.matrix(train_df[, feats]),
                       label = train_df$harmonized_value)
    dv  <- xgb.DMatrix(data = as.matrix(val_df[, feats]),
                       label = val_df$harmonized_value)

    fit <- xgb.train(params = params, data = dtr, nrounds = nrounds,
                     evals = list(train = dtr, val = dv),
                     verbose = 0,
                     early_stopping_rounds = early_stopping_rounds)

    predict(fit, dtest)
  })

  test_pred <- rowMeans(do.call(cbind, fold_preds))

  tibble(test_partition = test_partition,
         seed = seed,
         rmse = rmse(test$harmonized_value, test_pred),
         mae = mae(test$harmonized_value, test_pred),
         mape = mape(test$harmonized_value, test_pred),
         bias = bias(test$harmonized_value, test_pred),
         p.bias = percent_bias(test$harmonized_value, test_pred),
         smape = smape(test$harmonized_value, test_pred),
         r2 = cor(test$harmonized_value, test_pred)^2)
}
```

## Variability across the 5 HUC4-grouped splits

Rotate the held-out test partition across all 5 of script 03's partitions,
holding features/hyperparameters/seed fixed (`seed = 47`, matching 04's
default), so each of the 5 splits gets its own independent
leave-one-partition-out CV + test evaluation.

```{r split-repeat-run}
split_results <- map_dfr(1:5, ~ run_split(split_dataframes_ml, feats, final_hyperparams,
                                          test_partition = .x, seed = 47))

write_csv(split_results, "regional_clarity/xg_models/prelim/split_repeat_results.csv")

split_results
```

Summarize variability across the 5 held-out partitions - these are the
headline numbers for characterizing split-to-split variability.

```{r split-repeat-summary}
split_summary <- split_results %>%
  select(-seed) %>%
  pivot_longer(-test_partition, names_to = "metric", values_to = "value") %>%
  summarise(mean = mean(value),
            sd = sd(value),
            min = min(value),
            max = max(value),
            .by = metric)

split_summary
```

Plot each metric by held-out partition, to see whether error concentrates on
any one split rather than being evenly spread.

```{r split-repeat-plot}
split_results %>%
  select(-seed) %>%
  pivot_longer(-test_partition, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = factor(test_partition), y = value)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "held-out test partition", y = NULL) +
  theme_bw()
```

## Sensitivity to random seed

Fix the held-out test partition at 4 (04's original choice) and repeat the
same `run_split()` training/evaluation across many seed values, to see how
much test metrics move purely from XGBoost's internal stochastic elements
(`subsample`, `colsample_bytree` both < 1, which depend on R's RNG state
during `xgb.train`). `n_seeds` is deliberately small enough to run in a
reasonable amount of time - each seed costs about as much as one pass of
04's final-model-training step (4 fold fits) - increase it if a tighter
estimate of the variability is needed.

```{r seed-sim-run}
n_seeds <- 30

seed_results <- map_dfr(1:n_seeds, ~ run_split(split_dataframes_ml, feats, final_hyperparams,
                                                test_partition = 4, seed = .x))

write_csv(seed_results, "regional_clarity/xg_models/prelim/seed_sensitivity_results.csv")

seed_results
```

Summarize variability across seeds, including coefficient of variation
(sd / mean) so the spread can be judged relative to each metric's own scale.

```{r seed-sim-summary}
seed_summary <- seed_results %>%
  select(-test_partition) %>%
  pivot_longer(-seed, names_to = "metric", values_to = "value") %>%
  summarise(mean = mean(value),
            sd = sd(value),
            cv = sd(value) / mean(value),
            min = min(value),
            max = max(value),
            .by = metric)

seed_summary
```

Plot the distribution of each metric across seeds.

```{r seed-sim-plot}
seed_results %>%
  select(-test_partition) %>%
  pivot_longer(-seed, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15) +
  facet_wrap(~metric, scales = "free") +
  labs(x = NULL, y = "count (of 30 seeds)") +
  theme_bw()
```
