---
title: "Pull Weather Summaries"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 16, 2026"
output: html_document
# Quarto site build: show code only, don't re-run the fetches
execute:
  eval: false
---

**Supporting step · R · run from `regional_clarity/` · re-run only when the site set or date range changes; not executed when the site is built**

- **Purpose:** Fetch daily gridMET weather per site and summarize it into 1/3/7/30-day antecedent-weather features.
- **Inputs:** `aquamatch_files/filtered_regional_sdd.parquet`; gridMET.
- **Outputs:** `aquamatch_files/`: `weather_raw.feather`, `weather_summaries.feather`.
- **Runs after:** [02 · Make matchups](02_make_matches.Rmd) · **Used by:** [03 · Split data](03_split_data.ipynb)

This script pulls daily gridMET weather (precipitation, max/min/mean
temperature, solar radiation) for a set of AquaMatch sites over each site's
own observed date range, then summarizes it into "previous N days" rolling
antecedent-weather features for use as model features (joined in
`03_split_data.ipynb`).

It is **not** part of the numbered `00`–`07` pipeline: it doesn't need to be
re-run every time the pipeline runs, only when the active site set or date
range changes enough to be worth refreshing the cached lookup table it
writes. It started as `outlier_rework/21_gather_weather.R` +
`outlier_rework/22_weather_summaries.R` on the `outlier-removal-rework`
branch (a one-off research pull against that branch's own filtered dataset);
this version merges both stages into one script, generalized to run against
**this branch's** canonical AquaMatch outputs, so it's reproducible without
depending on another branch's files.

Weather comes from gridMET via `climateR::getGridMET()` — public, no auth
token needed (the alternative, NLDAS via Earthdata Login, needs a token this
project doesn't have configured). Each site is fetched individually over
just its own observed date range (plus a lag-window buffer), not the full
gridMET record, since most sites don't need more than that.

## Requirements for the input site table

Any AquaMatch-derived (or AquaMatch-shaped) table with one row per
observation and, at minimum, the columns below will work as
`sites_input_path`:

- `siteSR_id` — site identifier
- `lat`, `lon` — WGS84 coordinates
- `date` — satellite image date
- `harmonized_local_time` — field sample date/time (used only for its date)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(sf)
library(readr)
library(parallel)
library(climateR)
library(data.table)

log <- function(...) cat(sprintf("[%s] ", format(Sys.time(), "%H:%M:%S")), sprintf(...), "\n")

# -- input (AquaMatch-shaped; see "Requirements" above) --
# this branch's canonical filtered/matched site set (03_split_data.ipynb's
# own input, from 02_make_matches.Rmd)
sites_input_path <- "aquamatch_files/filtered_regional_sdd.parquet"

# -- outputs --
output_raw_path <- "aquamatch_files/weather_raw.feather"
output_summary_path <- "aquamatch_files/weather_summaries.feather"
output_failures_path <- "aquamatch_files/weather_fetch_failures.csv"

MAX_WINDOW_DAYS <- 30  # longest antecedent window summarized below
WINDOWS <- c(1, 3, 7, 30)
VARS <- c("pr", "tmmx", "tmmn", "srad")
N_WORKERS <- 8
```

## Per-site date ranges

The match window between satellite image and field SDD sample is +-5 days,
and which side leads is close to a 50/50 split across the dataset. "Previous
N days before the image date" only covers the image-to-field gap when the
field sample came *first* — when the image comes first, that window never
reaches forward to the field date at all, silently missing real antecedent
conditions for roughly half the dataset. Anchoring each site's fetch range on
`max(image_date, field_date)` (widened by `MAX_WINDOW_DAYS + 5` days on the
early end) avoids that, and matches `03_split_data.ipynb`'s own
`anchor_date` logic for joining these summaries back onto the modeling
dataset.

```{r site-ranges}
sites_raw <- read_parquet(sites_input_path,
                          col_select = c("siteSR_id", "lat", "lon", "date", "harmonized_local_time"))
sites_raw <- sites_raw %>%
  mutate(date = as.Date(date),
         field_date = as.Date(harmonized_local_time),
         anchor_date = pmax(date, field_date))

site_ranges <- sites_raw %>%
  group_by(siteSR_id, lat, lon) %>%
  summarize(min_date = min(date) - (MAX_WINDOW_DAYS + 5), max_date = max(anchor_date), .groups = "drop")
log("fetching gridMET for %s sites", nrow(site_ranges))
```

## Fetch gridMET

`mclapply` (fork-based parallelism) crashes on macOS here — `climateR`'s
curl/geospatial dependencies touch Objective-C runtime state that isn't
fork-safe ("may have been in progress in another thread when fork() was
called ... Crashing instead"). PSOCK workers (genuinely separate processes,
not forks) avoid that.

```{r fetch-function}
fetch_one <- function(i, site_ranges) {
  row <- site_ranges[i, ]
  fetch <- function() {
    pt <- sf::st_as_sf(data.frame(lon = row$lon, lat = row$lat),
                        coords = c("lon", "lat"), crs = 4326)
    res <- getGridMET(AOI = pt, varname = VARS,
                       startDate = as.character(row$min_date),
                       endDate = as.character(row$max_date))
    res$siteSR_id <- row$siteSR_id
    res
  }
  tryCatch(fetch(), error = function(e) {
    Sys.sleep(2)  # one retry after a short pause - transient network/server hiccups
    tryCatch(fetch(), error = function(e2) {
      data.frame(siteSR_id = row$siteSR_id, error = conditionMessage(e2))
    })
  })
}
```

```{r fetch-run}
t0 <- Sys.time()
cl <- makeCluster(N_WORKERS, type = "PSOCK")
clusterEvalQ(cl, { suppressMessages(library(climateR)); suppressMessages(library(sf)) })
clusterExport(cl, c("site_ranges", "VARS", "fetch_one"))
results <- parLapply(cl, seq_len(nrow(site_ranges)), fetch_one, site_ranges = site_ranges)
stopCluster(cl)
log("fetch took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

ok <- Filter(function(x) !"error" %in% names(x), results)
failed <- Filter(function(x) "error" %in% names(x), results)
log("succeeded: %s sites, failed: %s sites", length(ok), length(failed))

if (length(failed) > 0) {
  write_csv(bind_rows(failed), output_failures_path)
  log("failed site IDs written to %s", output_failures_path)
}

weather_raw <- bind_rows(ok)
log("total weather rows: %s", nrow(weather_raw))
write_feather(weather_raw, output_raw_path)
```

## Summarize into rolling antecedent-weather features

Total precip, max/mean/min temp, and total solar radiation over rolling
windows ending the day *before* each observation (`data.table::froll*` for
efficiency).

```{r summarize}
weather <- read_feather(output_raw_path)
log("loaded %s weather rows across %s sites", nrow(weather), n_distinct(weather$siteSR_id))

# units: gridMET pr is mm, tmmx/tmmn are Kelvin, srad is W/m^2
weather <- weather %>%
  mutate(date = as.Date(date),
         precip_mm = pr,
         tmax_degC = tmmx - 273.15,
         tmin_degC = tmmn - 273.15,
         tmean_degC = (tmax_degC + tmin_degC) / 2,
         srad_Wm2 = srad) %>%
  select(siteSR_id, date, precip_mm, tmax_degC, tmin_degC, tmean_degC, srad_Wm2)

setDT(weather)
setorder(weather, siteSR_id, date)

site_splits <- split(weather, by = "siteSR_id", keep.by = TRUE)

make_summary_one_site <- function(df) {
  site_id <- df$siteSR_id[1]
  # frollsum/frollmean/frollapply assume a regular series - gridMET
  # shouldn't have gaps, but guard anyway with one row per calendar day
  full_days <- data.table(date = seq(min(df$date), max(df$date), by = "1 day"))
  df <- merge(full_days, df[, -"siteSR_id"], by = "date", all.x = TRUE)
  df[, siteSR_id := site_id]

  for (w in WINDOWS) {
    df[[paste0("precip_mm_prev", w)]] <- frollsum(df$precip_mm, n = w)
    df[[paste0("tmax_degC_prev", w)]] <- frollapply(df$tmax_degC, N = w, FUN = max)
    df[[paste0("tmean_degC_prev", w)]] <- frollmean(df$tmean_degC, n = w)
    df[[paste0("tmin_degC_prev", w)]] <- frollapply(df$tmin_degC, N = w, FUN = min)
    df[[paste0("srad_Wm2_prev", w)]] <- frollmean(df$srad_Wm2, n = w)
  }
  # shift so window w on `date` covers the w days strictly BEFORE date
  df[, date := date + 1]
  df
}

log("computing rolling summaries across %s sites", length(site_splits))
t0 <- Sys.time()
summaries <- rbindlist(lapply(site_splits, make_summary_one_site), fill = TRUE)
log("took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

summaries <- summaries[!is.na(siteSR_id)]
```

## Export

```{r export}
write_feather(as.data.frame(summaries), output_summary_path)
log("wrote %s: %s rows, %s cols", output_summary_path, nrow(summaries), ncol(summaries))
```
