---
title: "Pull Site Characteristics"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 16, 2026"
output: html_document
---

This script pulls static site/catchment characteristics — point elevation and
LakeCat catchment land-cover metrics — for a set of AquaMatch sites, and
writes a small per-site lookup table for use as model features (joined in
`03_split_data.ipynb`).

It is **not** part of the numbered `00`–`05` pipeline: it doesn't need to be
re-run every time the pipeline runs, only when the active site set changes
enough to be worth refreshing the cached lookup table it writes. It started
as `outlier_rework/20_site_characteristics.R` on the `outlier-removal-rework`
branch (a one-off research pull against that branch's own filtered dataset);
this version is generalized to run against **this branch's** canonical
AquaMatch outputs, so it's reproducible without depending on another
branch's files.

Variable choice follows Topp et al. 2021 (*Environ. Res. Lett.* 16 055025),
which paired optical variables with LakeCat catchment metrics
(imperviousness, urban/forest/cropland/wetland land cover, all from the 2006
land-cover snapshot) for a national Secchi-depth model.

## Requirements for the input site table

Any AquaMatch-derived (or AquaMatch-shaped) table with one row per
observation and, at minimum, the columns below will work as
`sites_input_path` — this script `distinct()`s down to one row per site
before fetching anything:

- `siteSR_id` — site identifier
- `lat`, `lon` — WGS84 coordinates

`site_metadata_path` is AquaMatch's own raw site list (the
`siteSR_collated_WQP_NWIS_sites_with_NHD_info_*.csv` produced upstream of
this project, e.g. by `AquaMatch_siteSR_WQP`'s `add_NHD_waterbody_to_sites.R`
step) — it carries the NHDPlusV2 waterbody linkage (`wb_nhd_id`) each site
was already matched to, which is what LakeCat is keyed on.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(readr)
library(sf)
library(elevatr)
library(StreamCatTools)

log <- function(...) cat(sprintf("[%s] ", format(Sys.time(), "%H:%M:%S")), sprintf(...), "\n")

# -- inputs (AquaMatch-shaped; see "Requirements" above) --
# this branch's canonical filtered/matched site set (03_split_data.ipynb's
# own input, from 02_make_matches.Rmd)
sites_input_path <- "aquamatch_files/filtered_regional_sdd.parquet"
# AquaMatch's raw site metadata, carrying the NHDPlusV2 waterbody linkage
site_metadata_path <- "aquamatch_files/downloads/siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv"

# -- output --
output_path <- "aquamatch_files/site_characteristics.feather"
```

## Unique sites in the input set

```{r unique-sites}
final <- read_parquet(sites_input_path, col_select = c("siteSR_id", "lat", "lon"))
sites_unique <- distinct(final, siteSR_id, lat, lon)
log("n unique sites: %s", nrow(sites_unique))

sites_meta <- read_csv(site_metadata_path, show_col_types = FALSE) %>%
  select(siteSR_id, wb_nhd_id, wb_nhd_source, wb_gnis_name, wb_areasqkm)
sites_unique <- sites_unique %>% left_join(sites_meta, by = "siteSR_id")
log("sites with a valid wb_nhd_id (comid): %s / %s",
    sum(!is.na(sites_unique$wb_nhd_id)), nrow(sites_unique))
```

`wb_nhd_id` is treated below as an NHDPlusV2 comid — safe to join straight
into LakeCat with no crosswalk, *provided* every site's `wb_nhd_source` is
`"NHDPlusv2"` (the alternative, `Permanent_Identifier`, only fires for
non-CONUS HUC4s >= 1900). This was true for all 1,525 sites in the original
research pull; re-checked here every run rather than assumed, since the
active site set can change as `01`/`02` are reworked.

```{r verify-nhd-source}
stopifnot("some sites resolve to a non-NHDPlusv2 waterbody source - LakeCat join below is not safe as-is" =
            all(sites_unique$wb_nhd_source == "NHDPlusv2" | is.na(sites_unique$wb_nhd_source)))
```

## 1. Elevation (point, AWS terrain tiles)

```{r fetch-elevation}
log("fetching elevation for all sites")
pts_sf <- st_as_sf(sites_unique, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
t0 <- Sys.time()
elev <- get_elev_point(pts_sf, src = "aws", z = 9)
log("elevation fetch took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

elev_df <- st_drop_geometry(elev) %>% select(siteSR_id, elevation_m = elevation)
sites_unique <- sites_unique %>% left_join(elev_df, by = "siteSR_id")
log("elevation summary:"); print(summary(sites_unique$elevation_m))
```

## 2. LakeCat catchment characteristics

Land-cover-focused set, matching Topp et al.'s variable categories:
imperviousness, urban/forest/cropland/wetland land cover (2006, static, same
year Topp used), plus catchment area (LakeCat provides this for free
alongside any metric query).

```{r fetch-lakecat}
log("fetching LakeCat catchment metrics")
lakecat_metrics <- paste(c(
  "pctimp2006", "pcturbhi2006", "pcturblo2006", "pcturbmd2006", "pcturbop2006",
  "pctconif2006", "pctdecid2006", "pctmxfst2006",
  "pcthbwet2006", "pctwdwet2006", "pctcrop2006"
), collapse = ",")

comids <- unique(na.omit(sites_unique$wb_nhd_id))
log("n unique comids: %s", length(comids))

# batch in chunks to keep the request URL a reasonable length
chunk_size <- 200
chunks <- split(comids, ceiling(seq_along(comids) / chunk_size))
lakecat_raw <- purrr::map_dfr(chunks, function(ch) {
  lc_get_data(comid = paste(ch, collapse = ","), metric = lakecat_metrics,
              aoi = "catchment", showAreaSqKm = TRUE)
})
log("LakeCat rows returned: %s / %s comids requested", nrow(lakecat_raw), length(comids))

lakecat <- lakecat_raw %>%
  mutate(comid = as.character(comid),
         pct_forest_2006 = pctconif2006cat + pctdecid2006cat + pctmxfst2006cat,
         pct_wetland_2006 = pcthbwet2006cat + pctwdwet2006cat,
         pct_urban_2006 = pcturbhi2006cat + pcturblo2006cat + pcturbmd2006cat + pcturbop2006cat) %>%
  select(comid, catchment_area_sqkm = catareasqkm,
         pct_impervious_2006 = pctimp2006cat, pct_urban_2006, pct_forest_2006,
         pct_cropland_2006 = pctcrop2006cat, pct_wetland_2006)

sites_unique <- sites_unique %>% left_join(lakecat, by = c("wb_nhd_id" = "comid"))
log("sites with LakeCat data: %s / %s", sum(!is.na(sites_unique$pct_impervious_2006)), nrow(sites_unique))
```

## Export

```{r export}
write_feather(sites_unique %>% select(-wb_nhd_source), output_path)
log("wrote %s (%s cols, %s rows)", output_path, ncol(sites_unique) - 1, nrow(sites_unique))
```
